---
title: "Getting Started with geContour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with geContour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `geContour` package provides tools to visualize sparse spatial data as contour plots and export them as Google Earth overlays. This makes it easy to explore patterns in your geographic data within the Google Earth environment.

## Basic Usage

First, let's load the package:

```{r setup}
library(geContour)
```

### Creating a Simple Dataset

Let's create a sample dataset with latitude, longitude, and some values:

```{r}
# Generate sample data
set.seed(42)
sample_data <- data.frame(
  lat = runif(50, 30, 35),
  lon = runif(50, -120, -115),
  value = runif(50, 0, 100)
)

head(sample_data)
```

### Creating a Contour Plot

Now we can create a contour plot from this data:

```{r, fig.width=7, fig.height=5}
# Create a basic contour plot
contour_plot <- create_contour_plot(
  data = sample_data,
  interpolate = TRUE,
  interp_method = "akima"
)

# Display the plot
print(contour_plot)
```

### Exporting for Google Earth

To export this visualization for use in Google Earth, we would use:

```{r, eval=FALSE}
# Export to Google Earth (not actually run in this vignette)
result <- contour_to_google_earth(
  data = sample_data,
  output_name = "sample_overlay",
  alpha = 0.7,
  buffer_percent = 5
)

# The result contains paths to the created files
result$png  # Path to PNG image
result$kml  # Path to KML file
result$kmz  # Path to KMZ file
```

## Customization Options

The package provides several options for customizing your visualizations:

### Custom Colors

```{r, fig.width=7, fig.height=5}
# Using a custom color palette
custom_colors <- colorRampPalette(c("blue", "cyan", "yellow", "red"))(10)

custom_plot <- create_contour_plot(
  data = sample_data,
  fill_colors = custom_colors,
  alpha = 0.8
)

print(custom_plot)
```

### Custom Contour Breaks

```{r, fig.width=7, fig.height=5}
# Using a specific number of contour levels
break_plot <- create_contour_plot(
  data = sample_data,
  contour_breaks = 5
)

print(break_plot)
```

## Conclusion

The `geContour` package makes it easy to visualize spatial data and export it for use in Google Earth. For more details, refer to the function documentation using `?contour_to_google_earth`.br