# Examples

The examples show some challenges, work-around, and exploitation of the features of both the `tiny_pedigree` function and `gt` tables.

Initialize the needed packages.

```{r}
#| label: core

## Activate the Core Packages
library(tidyverse) ## Brings in a core of useful functions
library(gt)        ## Tables

## special libraries
library(kinship2)  ## Core package to calculate and plot
library(R.devices) ## External plot files (e.g., PNG)

source("Functions.R", local=TRUE)

```

## Eight Generations

There are a lot of individuals to fit on this pedigree diagram. Perhaps too many as the typesize, in particular, must be very small.

Note that there is an interaction between the `typesize` and `fold` parameters. Folding the names doesn't help.

Warning: row (generation) spacing seems to be disrupted at the lowest row in a many-generation diagram if there are several individuals at the lowest level.

```{r}


data <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
 "ID, dad, mom, gender, born, died
 Charles Wentworth, NA, NA, male, 1648, 1780
 Amariah Wentworth, Charles Wentworth, Bethiah Fenno, male, 1719, 1799
 Bethiah Fenno, NA, NA, female, 1692, 1780 
 Rebecca Shepard, NA, NA, female, 1723, 1791
 Elijah Wentworth, Amariah Wentworth, Rebecca Shepard, male, 1775, 1810
 Elijah S. Estes, NA, NA, male, 1814, 1887
 Zebiah Walker Wentworth, Elijah Wentworth, Lucy Walker, female, 1810, 1887
 Lucy Walker, NA, NA, female, NA, NA
 Mary Adeline Palmer, NA, NA, female, 1848, 1918
 Lorenzo Langston Estes, Elijah S. Estes, Zebiah Walker Wentworth, male, 1839, 1909
 Addie Estes, Lorenzo Langston Estes, Mary Adeline Palmer, female, NA, NA
 Charles A. Bridges, NA, NA, male, 1877, NA
 Charles L. Bridges, Charles A. Bridges, Addie Estes, male, 1912, 2001
 Mary Jane Bridges, NA, NA, female, 1912, 2016
 Thomas Bridges, Charles L. Bridges, Mary Jane Bridges, male, 1938, NA
 Kent Bridges, Charles L. Bridges, Mary Jane Bridges, male, 1941, NA")

## Assign or calculate enhancements
data$hilite <- TRUE
data <- data |>
 mutate(color = case_when(
 gender == "female" ~ "lightblue",
 gender == "male" ~ "lightsalmon1"))

## Create a presentation table
gt(data) |>
 cols_hide(columns = c(dad, mom, gender, hilite, color)) |>
 tab_source_note(source_note="Source: Bridges family documents")

tiny_pedigree(data=data, textsize=0.5, symbolsize=0.8, fold=FALSE)

```

Breaking the pedigree into smaller sets is a way around some of the limitations seen in the last pedigree diagram.

```{r}
#| label: twoparts

## Divide the data into two halves

## First Half
data1 <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
 "ID, dad, mom, gender, born, died
 Charles Wentworth, NA, NA, male, 1648, 1780
 Amariah Wentworth, Charles Wentworth, Bethiah Fenno, male, 1719, 1799
 Bethiah Fenno, NA, NA, female, 1692, 1780 
 Rebecca Shepard, NA, NA, female, 1723, 1791
 Elijah Wentworth, Amariah Wentworth, Rebecca Shepard, male, 1775, 1810
 Elijah S. Estes, NA, NA, male, 1814, 1887
 Zebiah Walker Wentworth, Elijah Wentworth, Lucy Walker, female, 1810, 1887
 Lucy Walker, NA, NA, female, NA, NA
 Lorenzo Langston Estes, Elijah S. Estes, Zebiah Walker Wentworth, male, 1839, 1909")

## Second Half
data2 <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
 "ID, dad, mom, gender, born, died
 Mary Adeline Palmer, NA, NA, female, 1848, 1918
 Lorenzo Langston Estes, NA, NA, male, 1839, 1909
 Addie Estes, Lorenzo Langston Estes, Mary Adeline Palmer, female, NA, NA
 Charles A. Bridges, NA, NA, male, 1877, NA
 Charles L. Bridges, Charles A. Bridges, Addie Estes, male, 1912, 2001
 Mary Jane Bridges, NA, NA, female, 1912, 2016
 Thomas Bridges, Charles L. Bridges, Mary Jane Bridges, male, 1938, NA
 Kent Bridges, Charles L. Bridges, Mary Jane Bridges, male, 1941, NA")

## Assign or calculate enhancements to the first half
data1$hilite <- TRUE
data1 <- data1 |>
 mutate(color = case_when(
 gender == "female" ~ "lightblue",
 gender == "male" ~ "lightsalmon1"))

## Generate a pedigree from the first half
tiny_pedigree(data=data1, symbolsize=1.1, fold=FALSE)

## Assign or calculate enhancements to the second half
data2$hilite <- TRUE
data2 <- data2 |>
 mutate(color = case_when(
 gender == "female" ~ "lightblue",
 gender == "male" ~ "lightsalmon1"))

## Generate a pedigree from the second half
tiny_pedigree(data=data2, textsize=0.6, symbolsize=1.1, fold=FALSE)

## Join (row bind) the two halves
## The first two lines of second half are not needed
data <- rbind(data1, data2[3:nrow(data2),])

## Create a presentation table
gt(data) |>
 cols_hide(columns=c(dad, mom, gender, hilite, color)) |>
 tab_source_note(source_note="Source: Bridges family documents")

```

## Interpretive information

You can use the color coding to provide information that is ***not*** related to the demographic properties of the data.

In the following example, research progress is noted on the pedigree diagram with a color code (green means already interviewed).

```{r}
#| label: very-simple

## Read the basic data (all people are listed)
data <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
 "ID, dad, mom, gender, interviewed 
 Genji, NA, NA, male, no 
 Shizuko, NA, NA, female, yes
 Hiroshi, Genji, Shizuko, male, no 
 Takashi, Genji, Shizuko, male, yes 
 Osamu, Genji, Shizuko, male, yes 
 Akari, NA, NA, female, no 
 Fumiko, Osamu, Akari, female, no
 Yui, NA, NA, female, yes
 Shota, Hiroshi, Yui, male, no
 Sakura, NA, NA, female, yes")

## Read links for people without children
links <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
  "id1, id2
  Takashi, Sakura")

## Print a presenttion table
gt(data) |>
 sub_missing(missing_text = "") |>
 cols_hide(columns=c(gender)) |>
 tab_source_note(source_note="Source: Example data")

## Calculate and assign column properties
data <- data |>
 mutate(color = case_when(
 interviewed == "yes" ~ "lightgreen",
 interviewed == "no" ~ "gray85")) |>
 mutate(hilite = TRUE)

## Generate and plot the pedigree structure 
## You need to use the link parameter for "no-children" couples
tiny_pedigree(data=data, links=links)

```

## Family with lots of children

This is an example of the complexity when there are quite a few children. The software seems to handle this number of individuals, but it is likely close to the limit.

```{r}
data <- read_csv(col_names = TRUE, show_col_types=FALSE, file=
 "ID, father, mother, sex, status
 Howard, NA, NA, male, 1
 Viola, NA, NA, female, 1
 William, Howard, Viola, male, 0
 Alice, Howard, Viola, female, 1
 Wesley, Howard, Viola, male, 0
 David, Howard, Viola, male, 0
 Nancy, Howard, Viola, female, 0 
 Kimo, Howard, Viola, male, 0
 Edward, Howard, Viola, male, 0
 Claudia, NA, NA, female, 0
 Karen, NA, NA, female, 1
 Ayesha, NA, NA, female, 0
 Laurie, NA, NA, female, 0
 Michael, NA, NA, male, 0
 Michiko, NA, NA, female, 0
 Linda, NA, NA, female, 0
 Kim, NA, NA, male, 0
 Angel, William, Claudia, female, 0
 Jolene, William, Claudia, female, 0
 Pono, Wesley, Karen, male, 0
 Gabriel, Wesley, Ayesha, male, 0
 Sokah, Wesley, Laurie, female, 0
 Jessie, Edward, Michiko, male, 0
 Nalani, Kimo, Linda, female, 0
 Leilani, Michael, Alice, female, 0
 Sefu, Michael, Alice, male, 0
 Eric, NA, NA, male, 0
 Scott, NA, NA, male, 0
 Shane, Scott, Angel, male, 0
 Skyler, Scott, Angel, female, 0
 Kalani, NA, NA, male, 0
 Dysen, Kalani, Jolene, male, 0
 Jordyn, Kalani, Jolene, female, 0
 Laura, NA, NA, female, 0
 Salem, Sefu, Laura, male, 0
 Jennifer, NA, NA, female, 0
 Christopher, Gabriel, Jennifer, male, 0
 Vik, NA, NA, female, 0")

## Links for individuals without children
links <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
  "id1, id2
  Kim, Nancy
  Eric, Leilani
  Jessie, Vik")

## Generate a presentation table
gt(data) |>
 cols_hide(columns=c(sex, status)) |>
 sub_missing(missing_text = "") |>
 tab_source_note(source_note="Source: Furumoto family data")

## Generate the pedigree data structure (with no-children couples)
tiny_pedigree(data=data, links=links, 
 symbolsize=0.8, textsize = 0.4, fold=FALSE)

```

## A complex family

```{r}

## Read the data
data <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
 "ID, dad, mom, gender, dead
 Yuji Sorakubo, NA, NA, male, 1
 Kunichi Furumoto, NA, NA, male, 1
 Kiyo Sorakubo, NA, NA, female, 1
 Ayako Sorakubo, Yuji Sorakubo, Kiyo Sorakubo, female, 1
 Satayoshi Furumoto, Yuji Sorakubo, Kiyo Sorakubo, male, 1
 Shigeo Sorakubo, Yuji Sorakubo, Kiyo Sorakubo, male, 1
 Howard Furumoto, Kunichi Furumoto, Kiyo Sorakubo, male, 1
 Hirui Furumoto, Kunichi Furumoto, Kiyo Sorakubo, female, 1 
 Hisano Fujiwara, Kunichi Furumoto, Kiyo Sorakubo, female, 1
 George Fujiwara, NA, NA, male, 1
 Mom Sorakubo, NA, NA, female, 1
 Carol Sorakubo, NA, NA, female, 0
 Clyde Sorakubo, Satayoshi Furumoto, Mom Sorakubo, male, 1
 Jane Sorakubo, Satayoshi Furumoto, Mom Sorakubo, female, 0
 Karen Sorakubo, Clyde Sorakubo, Carol Sorakubo, female, 0
 Amy Sorakubo, Clyde Sorakubo, Carol Sorakubo, female, 1 
 Kiyoko Sorakubo, NA, NA, female, 1
 Michael Sorakubo, Shigeo Sorakubo, Kiyoko Sorakubo, male, 0
 Wendy Sorakubo, NA, NA, female, 0
 Myra Yamachi, George Fujiwara, Hisano Fujiwara, male, 0
 David Yamachi, NA, NA, male, 1")

## Links for the individuals without children
links <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
  "id1, id2
  Michael Sorakubo, Wendy Sorakubo
  Myra Yamachi, David Yamachi")

## Create a presentation table
gt(data) |>
 cols_hide(columns=c(gender, dead)) |>
 sub_missing(missing_text = "") |>
 tab_source_note(source_note="Source: Furumoto family data")

## Color all the symbols 
data$color <- "gray85"
data$hilite <- TRUE

## Modify the color of a specific row
data$color[7] <- "pink"

## Generate the plot the pedigree
tiny_pedigree(data=data, links=links,
              textsize = 0.7,
              symbolsize = 0.8)

```

## Two families combined

```{r}

## Read the data
data <- read_csv(col_names = TRUE, show_col_types=FALSE, file=
 "ID, father, mother, sex, status
 Herbert Kawahara, NA, NA, male, 1
 Hideko Kawahara, NA, NA, female, 1
 Brian Kawahara, Herbert Kawahara, Hideko Kawahara, male, 0
 Nancy Kawahara, NA, NA, female, 0
 Candy Kawahara, Herbert Kawahara, Hideko Kawahara, female, 0
 Debbie Kawahara, Herbert Kawahara, Hideko Kawahara, female, 0
 Neil Nagata, NA, NA, male, 0
 Doreen Kawahara, Herbert Kawahara, Hideko Kawahara, female, 0
 Ayako Sorakubo, NA, NA, female, 1
 Jinpachi Nakashima, NA, NA, male, 1
 Colleen Kawahara, Herbert Kawahara, Hideko Kawahara, female, 0
 Kevin Nakashima, Jason Nakashima, Eleanor Oyama, male, 0
 Terry Nakashima, Jason Nakashima, Eleanor Oyama, male, 0
 Cory Nakashima, Terry Nakashima, Colleen Kawahara, male, 0
 Karen Nakashima, Terry Nakashima, Colleen Kawahara, female, 0
 Jason Nakashima, Jinpachi Nakashima, Ayako Sorakubo, male, 1
 Sachiko Nakashima,  Jinpachi Nakashima, Ayako Sorakubo, female, 1
 Jerry Nakashima, Jinpachi Nakashima, Ayako Sorakubo, male, 1
 Eleanor Oyama, NA, NA, female, 1
 Sharon Nakashima, Jason Nakashima, Eleanor Oyama, female, 0
 Alyssa Denison, David Denison, Sharon Nakashima, female, 0
 Kellianne Denison, David Denison, Sharon Nakashima, female, 0
 Christopher Denison, David Denison, Sharon Nakashima, male, 0
 David Denison, NA, NA, male, 0")

## Links for married individuals with no children
links <- read_csv(col_names=TRUE, show_col_types=FALSE, file=
  "id1, id2
  Brian Kawahara, Nancy Kawahara
  Neil Nagata, Debbie Kawahara")

## Create a presentation table
gt(data) |>
 cols_hide(columns=c(sex, status)) |>
 sub_missing(missing_text = "") |>
 tab_source_note(source_note="Source: Kawahara family data")

## Generate and plot the pedigree diagram
tiny_pedigree(data = data, links = links,
              symbolsize = 0.8,
              textsize = 0.6)
```
