---
title: "Voronoi Tessellation Analysis — Phase 1"
subtitle: "Sensitivity Testing with Simulated Data"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
    fig_caption: true
    keep_tex: false
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{booktabs}
params:
  results_file: "phase1_results.rds"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(kableExtra)

all_results <- readRDS(params$results_file)
source("voronoi_functions.R")
```

# Introduction

This report presents the results of a Voronoi tessellation analysis
using simulated data. **Anchor** points define Voronoi cells that
partition the study area. **Sample** points are distributed across the
area under various patterns (random, clustered, or mixed). For each
cell, a Poisson test evaluates whether the observed number of samples
differs significantly from the number expected under a uniform null
hypothesis. A chi-squared goodness-of-fit test provides an overall
assessment.

The goal is to understand how different spatial patterns of sample
points are detected by this methodology, building intuition for
interpreting real-world data in Phase 2.

------------------------------------------------------------------------

```{r run-sections, results='asis'}
for (i in seq_along(all_results)) {
  current_run   <- all_results[[i]]
  current_index <- i
  cat(knitr::knit_child("_run_section.Rmd", envir = environment(), quiet = TRUE))
}
```

# Summary Across Runs

```{r summary-table}
summary_df <- bind_rows(lapply(seq_along(all_results), function(i) {
  r   <- all_results[[i]]
  cs  <- r$stats$cell_stats %>% st_drop_geometry()
  chi <- r$stats$chi_squared

  tibble(
    Run       = i,
    Label     = r$params$label,
    Anchors   = r$params$n_anchors,
    Samples   = r$params$n_samples,
    Mode      = r$params$sample_mode,
    `Sig. Cells`  = sum(cs$significant),
    `Total Cells` = nrow(cs),
    `Chi² p`  = chi$p.value
  )
}))

kable(summary_df, format = "latex", booktabs = TRUE, linesep = "",
      digits = c(0, NA, 0, 0, NA, 0, 0, 4)) %>%
  kable_styling(position = "center", latex_options = c("hold_position", "scale_down"),
                full_width = FALSE)
```

```{r summary-interpretation, results='asis'}
n_runs      <- length(all_results)
any_sig     <- any(sapply(all_results, function(r) any(r$stats$cell_stats$significant)))
clustered   <- sapply(all_results, function(r) r$params$sample_mode != "random")
sig_counts  <- sapply(all_results, function(r) sum(r$stats$cell_stats$significant))

cat("\n\n")
if (any_sig) {
  cat(sprintf("Across %d runs, significant cell-level departures were detected in %d run(s). ",
              n_runs, sum(sig_counts > 0)))
  if (any(clustered & sig_counts > 0)) {
    cat("As expected, runs with clustered sample distributions tended to produce more significant cells, ")
    cat("confirming that the methodology can detect spatial clustering.")
  }
} else {
  cat(sprintf("Across all %d runs, no significant cell-level departures were detected, ",
              n_runs))
  cat("consistent with the null hypothesis of uniform sample distribution.")
}
cat("\n")
```
