# What a Voronoi Map Is

Think about where you buy groceries. If the markets are all about the same, you go to the nearest one. Draw a line around everyone who shops at each market and you've divided the town into districts. Every household inside a district is closer to its own market than to any other.

That's a Voronoi tessellation. Each point, here a market, gets a tile. The tile holds every place that's closer to its point than to any other point. The tiles fit together with no gaps and no overlaps, and together they cover the whole area.

This chapter builds one twice. First with five imaginary markets, small enough to see how it works. Then with twelve vegetation samples at Kīpuka Puaulu, which gives the map that opened Chapter 1.

```{r}
#| label: setupwhat
#| message: false

## --- Standard packages ---

## many useful functions, including read_csv() and ggplot()
library(tidyverse)
## formatted tables
library(gt)

## --- Specialized packages ---

## spatial data: points, polygons, and the Voronoi function
library(sf)

## --- Options and defaults ---

## suppress read_csv() column-type messages
options(readr.show_col_types = FALSE)

## the vegetation colors used throughout the document (Okabe-Ito palette)
veg_colors <- c("Kīpuka forest"     = "#009E73",
                "Koa-ʻōhiʻa forest" = "#0072B2",
                "ʻŌhiʻa forest"     = "#56B4E9",
                "ʻŌhiʻa scrub"      = "#E69F00",
                "Grass and savanna" = "#F0E442")
```

## Five markets

Here are five markets in a square town, ten kilometers on a side. The coordinates are in kilometers from the town's southwest corner.

```{r}
#| label: markets

## five markets, their positions in kilometers
markets <- read_csv(file =
  "market, easting, northing
   A,      2.0,     8.0
   B,      7.0,     8.5
   C,      5.0,     5.0
   D,      1.5,     2.0
   E,      8.0,     2.5")

## show the data as a table
markets |>
  gt() |>
  tab_source_note(source_note = "Imaginary markets, for demonstration.")
```

Three steps turn the points into tiles. The first builds the tiles. The second splits the result into one polygon per market. The third trims the tiles to the edge of town, because a tile on the outside of the group would otherwise run on forever.

```{r}
#| label: markettiles
#| fig-cap: "Five markets and their tiles. Every place inside a tile is closer to that tile's market than to any other."

## turn the table into spatial points
market_points <- st_as_sf(markets, coords = c("easting", "northing"))

## the town: a square, 10 km on a side
town_box <- st_bbox(c(xmin = 0, ymin = 0, xmax = 10, ymax = 10))
town     <- st_as_sfc(town_box)

## step 1: build the tiles around all the points at once
tiles_all <- st_voronoi(st_union(market_points), envelope = town)

## step 2: split the result into separate polygons
market_tiles <- st_collection_extract(tiles_all)

## step 3: trim the tiles to the town's edge
market_tiles <- st_intersection(market_tiles, town)

## the tiles come back in their own order, so attach each tile to the market inside it
market_tiles <- st_sf(geometry = market_tiles)
market_tiles <- st_join(market_tiles, market_points)

## draw the tiles, the markets, and their labels
ggplot() +
  geom_sf(data = market_tiles, fill = "grey92", color = "grey30") +
  geom_sf(data = market_points, size = 2.5) +
  geom_sf_text(data = market_points, aes(label = market), nudge_y = 0.45, size = 4.5) +
  coord_sf(expand = FALSE) +
  theme_minimal(base_size = 13) +
  labs(x = NULL, y = NULL)
```

## Halfway

Look at the edge between markets A and C. Every place on it is exactly as far from A as from C. That makes it the line that cuts the path between them in half, at a right angle. Each edge in the map is built the same way, from one pair of neighboring points.

The midpoint between A and C is easy to find: average their coordinates.

```{r}
#| label: halfway
#| fig-cap: "The edge between A and C crosses the path between them (dashed) at its midpoint (open circle), at a right angle."

## the two markets
pair <- filter(markets, market %in% c("A", "C"))

## the midpoint of the path between them
midpoint <- tibble(easting  = mean(pair$easting),
                   northing = mean(pair$northing))

## show the midpoint as a labelled table
midpoint |>
  gt() |>
  tab_header(title = "Midpoint between markets A and C (km)")

## the path from A to C, as a one-row table of start and end points
path <- tibble(x    = pair$easting[1], y    = pair$northing[1],
               xend = pair$easting[2], yend = pair$northing[2])

## draw the tiles again, with the path from A to C and its midpoint
ggplot() +
  geom_sf(data = market_tiles, fill = "grey92", color = "grey30") +
  geom_segment(data = path, aes(x = x, y = y, xend = xend, yend = yend),
               linetype = "dashed") +
  geom_point(data = midpoint, aes(easting, northing), shape = 21, fill = "white", size = 3.5) +
  geom_sf(data = market_points, size = 2.5) +
  geom_sf_text(data = market_points, aes(label = market), nudge_y = 0.45, size = 4.5) +
  coord_sf(expand = FALSE) +
  theme_minimal(base_size = 13) +
  labs(x = NULL, y = NULL)
```

So a Voronoi map makes one assumption about the ground between the points. **Whatever is true at a point is true halfway to its neighbors.** Past the halfway line, the neighbor takes over.

## What the map uses, and what it doesn't

A Voronoi map uses three things and nothing else:

- where the points are,
- what category each point belongs to, and
- the border of the study area.

It doesn't know about roads, rivers, ridges or lava flows. If a river runs between two markets, the tile edge crosses it anyway. For markets, that might be a real weakness. People don't shop across a river with no bridge.

The method has several names, because it was found more than once. Mathematicians know it from the work of Peter Gustav Lejeune Dirichlet in 1850 and Georgy Voronoi in 1908. Geographers and hydrologists often say Thiessen polygons, after Alfred Thiessen, who used them in 1911 to average rainfall from a scatter of rain gauges. It's the same construction under each name.

<!-- VERIFY before publishing: Dirichlet 1850, Voronoi 1908, Thiessen 1911 (Monthly Weather Review), per the verify list. -->

It's fair to call the result *reproducible*. It's not quite fair to call it *objective*. Someone chose where the points went and where the border runs, and those choices shape every tile. But anyone given the same points and the same border gets exactly the same map. That's what a hand-drawn map can't promise.

## Twelve points at Kīpuka Puaulu

Now the same steps with real ground. The twelve sample points below are spread across a 3.2 by 2.6 kilometer rectangle around Kīpuka Puaulu. I placed them for this demonstration, three in each of four kinds of vegetation structure, and each one takes the vegetation type that the 1974 map shows at that spot. Later chapters replace them with real relevés.

The coordinates are in meters, in the UTM zone 5 north system that covers the island of Hawaiʻi.

```{r}
#| label: releves

## the twelve sample points
releves <- read_csv("data/kipuka_puaulu_releves.csv")

## show them
releves |>
  gt() |>
  tab_source_note(source_note = "Points placed for demonstration; vegetation types from Mueller-Dombois and Fosberg (1974).")

## the study area, given by its southwest and northeast corners
area_corners <- read_csv("data/kipuka_puaulu_study_area.csv")

## show it
area_corners |>
  gt() |>
  tab_header(title = "Study area corners (UTM zone 5N, meters)")
```

The steps are the ones used for the markets. The only change is that the points now carry a coordinate system, EPSG code 32605, so distances come out in meters.

```{r}
#| label: relevetiles
#| fig-cap: "The twelve points and their tiles, colored by vegetation type. This is the map that opens Chapter 1."
#| fig-height: 7.5

## turn the table into spatial points in UTM zone 5N
releve_points <- st_as_sf(releves, coords = c("easting", "northing"), crs = 32605)

## the study area as a rectangle
study_box <- st_bbox(c(xmin = area_corners$easting[1],
                       ymin = area_corners$northing[1],
                       xmax = area_corners$easting[2],
                       ymax = area_corners$northing[2]),
                     crs = st_crs(32605))
study_area <- st_as_sfc(study_box)

## step 1: build the tiles
tiles_all <- st_voronoi(st_union(releve_points), envelope = study_area)

## step 2: split into separate polygons
releve_tiles <- st_collection_extract(tiles_all)

## step 3: trim to the study area
releve_tiles <- st_intersection(releve_tiles, study_area)

## attach each tile to the point inside it
releve_tiles <- st_sf(geometry = releve_tiles)
releve_tiles <- st_join(releve_tiles, releve_points)

## draw the map
ggplot() +
  geom_sf(data = releve_tiles, aes(fill = vegetation), color = "white", linewidth = 0.6) +
  geom_sf(data = releve_points, shape = 21, fill = "white", color = "black", size = 2.4) +
  scale_fill_manual(values = veg_colors, breaks = names(veg_colors), name = NULL) +
  coord_sf(datum = NA, expand = FALSE) +
  theme_void(base_size = 13) +
  theme(legend.position = "bottom")
```

A map made of tiles also gives numbers. Each tile has an area, so each vegetation type does too.

```{r}
#| label: releveareas

## the area of each tile, converted from square meters to hectares
releve_tiles <- mutate(releve_tiles, area_ha = as.numeric(st_area(geometry)) / 10000)

## drop the geometry and add up the area by vegetation type
area_by_type <- st_drop_geometry(releve_tiles)
area_by_type <- group_by(area_by_type, vegetation)
area_by_type <- summarise(area_by_type, tiles = n(), area_ha = round(sum(area_ha), 1))
area_by_type <- arrange(area_by_type, desc(area_ha))

## show the result
area_by_type |>
  gt() |>
  tab_header(title = "Area by vegetation type, from the tiles")
```

Those numbers are only as good as the assumption behind them. Twelve points can't know what lies between them. The halfway rule fills that ground in, whether the vegetation agrees or not.

## Where the assumption fits

For markets and schools, the halfway rule is close to how people behave. You shop at the nearest market and your children go to the nearest school, most of the time.

For soils, Arnold and Milne found that it matched what the cartographer already did. With a sample every 50 to 200 meters, a hand-drawn soil map is close to a Voronoi map anyway.

Vegetation is harder. Relevés are expensive, so they're far apart. And the vegetation scientist drawing the map doesn't draw halfway lines. The line goes where the edge of the forest shows on the air photo. That's the gap between the two maps in Chapter 1, and it's why the question there is a real one.

::: {.callout-note}
## Useful?

**Where the points stand for their surroundings,** as with markets, schools or dense soil samples, yes. The map takes seconds and anyone can reproduce it.

**Where the points are sparse and the ground between them varies,** as with vegetation, not yet. The tiles know nothing about the land between the points. Later chapters work on that.
:::

The tiles know where the points are. They don't know anything else.
