# What the driving cost {#sec-driving-cost}

```{r}
#| label: setup
#| include: false

source("R/book_setup.R")
source("R/driving_cost_data.R")

D <- driving_cost_data()
P <- D$prairie
T3 <- D$three
CR <- D$cranes

stopifnot(nrow(P) == 3, nrow(T3) == 3, D$cream_gap > 0)
```

## The worst miles of the whole thing

The hardest driving I did in sixteen years was four days across the Canadian prairie in June of 2014, and I can tell you exactly why.

**We were plagued by potholes.** We often slowed to avoid damage, which in an 1,800-pound car with small wheels is not a precaution, it's a requirement. And slowing meant looking in the rear-view mirror, because there was always something large behind me that hadn't slowed. Hours of it. Those miles were very stressful.

Now here's what the record says about them.

```{r}
#| label: tbl-prairie
#| tbl-cap: "The prairie days. Gross speed is the whole day divided by the whole clock, stops and all, which is the only speed this log can produce."
#| echo: false
z <- P[, c("date","start_city","end_city","odo_miles","elapsed_hours","gross_mph")]
z$date <- format(z$date, "%d %b %Y")
names(z) <- c("Date", "From", "To", "Miles", "Hours", "Gross mph")
knitr::kable(z, align = c("l","l","l","r","r","r"))
```

**A median of `r D$prairie_med` miles an hour, against `r D$med_gross` for the whole record.** The worst days I can remember are among the *faster* days I ever drove.

I want to be careful here, because there's an obvious wrong conclusion sitting there. It's that my memory has inflated the whole thing. It hasn't.

## The correction that turns it around

When I first saw those numbers I said something that fixed the problem in one sentence. **I couldn't go slow, because the big vehicles kept moving at highway speeds.**

Read that again against the table. **The speed isn't evidence against the stress. The speed is the stress.** Slowing for a pothole while something enormous closes on you at seventy is precisely the situation I'm describing, and the only way it shows up in a two-point measurement is as a day that averaged out fast.

So my account doesn't contradict the record. **It predicts it.** Any honest version of that day, written into this log, comes out looking like an unremarkable fast run, and that's what it did.

## Three days, one shape

Once you've seen it once you start finding it everywhere, and by now this book has three of them.

```{r}
#| label: tbl-three
#| tbl-cap: "Three days I remember for three completely different reasons, and what the record made of each. Nothing in this table is wrong."
#| echo: false
z <- T3[, c("what","date","start_city","end_city","odo_miles","gross_mph")]
z$date <- format(z$date, "%d %b %Y")
names(z) <- c("What I remember", "Date", "From", "To", "Miles", "Gross mph")
knitr::kable(z, align = c("l","l","l","l","r","r"))
```

The first is the prairie, and it was fear. The second is the tree tunnel from @sec-roads, four hours of an excellent highway I couldn't wait to get off, and it was boredom. The third is @sec-long-days's frontage road, where I left the interstate on purpose to crawl past Nebraska farms, and it was pleasure.

**Terror, tedium and enjoyment. All three come out at or above the median.**

Vigilance has no column in this record. Neither does tedium. And neither, it turns out, does an afternoon spent looking at things.

## Why: the log samples the day twice

The reason isn't mysterious and it isn't a fault. It's arithmetic, and it was designed in.

**A trip log records where I started, when I started, where I stopped and when I stopped.** Two points, ten hours apart. Everything between them, every pothole and every truck and every decision to turn off onto a smaller road, is averaged into a single number by construction.

Gross speed is a two-point measurement of a ten-hour process. It can't report variation, because it never sampled any. **A day that ran steadily at eighty and a day that alternated between forty and seventy give you the same figure, and it's the second one that frightened me.**

That isn't the log failing. It recorded what it was built to record, faithfully, for six years. **What it never claimed to hold is what the day was like.**

## And the second car proves it

All of that would be an argument rather than a finding, except that a later car in this fleet kept a column the first one didn't.

**Creamsicle's sheets record stopped time.** That makes it possible to work out two speeds for the same day: the gross figure, which is all TwoRed can ever give us, and a net figure with the stopped hours taken out.

Across `r D$n_cream` timed legs, **the median gross speed is `r D$cream_gross` miles an hour and the median net speed is `r D$cream_net`.** A gap of **`r D$cream_gap`**.

That number is worth sitting with. It means **every gross speed printed anywhere in this book is a ceiling, roughly seven or eight miles an hour below the speed the car was actually travelling** when it was travelling. And it means the averaging I've been describing isn't a hypothesis. It's got a size.

The extreme case is my favourite thing in either record. `r format(as.Date(CR$date), "%d %B %Y")`, `r CR$lv_city` to `r CR$ar_city`. **`r CR$miles` miles. `r CR$raw_hours` hours, `r CR$stop_hours` of them stopped.** Gross speed `r CR$gross_mph` miles an hour, which is walking pace for a car.

The comment column on that row contains one word.

> `r CR$comment`

**A fifty-mile day that took five hours because we stopped to watch birds**, and the only reason anybody will ever know that is that a later log had somewhere to put it. Enter that same day in TwoRed's sheet and you get ten miles an hour and no explanation at all.

## What this licenses

I've spent this chapter arguing that the record can't see the driver, and it's fair to ask why a book largely made of that record would want to press the case so hard.

Here's why. **Everything after this point is analysis**, and analysis is where a record like this one becomes dangerous. Sixteen years of numbers, audited and corrected and cross-checked, give you answers that look authoritative because they're precise. The temptation is to let precision stand in for completeness.

So it's worth settling, before any of that starts, exactly what these instruments are. **A fuel log is a list of transactions. A trip log is a pair of timestamps.** Between them they'll reconstruct a continent, recover a mountain, date a crossing to the day and catch an error made with cold hands in 2012. They can't tell you that the driver was frightened, or bored, or delighted.

**They were never asked to.** The chapters that follow are honest about their limits because this one drew them.

## The thing that was never in the column

There's a version of this book that could have been written from the numbers alone. It'd be accurate. Every figure in it would survive an audit.

And it would have a hole in the middle of it exactly the shape of a person: one driver, alone for most of it, deciding when to slow down and when to push on, when to leave the interstate and when to endure it, and whether today was the day to try for the mountain again.

**The record kept the miles. I kept the rest.** That the two can still be laid alongside each other, fifteen years later, is the only reason any of this could be written down at all.
