---
title: "Photo File Documentation"
author: "K. W. Bridges"
date: "8/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}

## Set some useful chunk options
knitr::opts_chunk$set(echo    = FALSE)  ## FALSE suppresses the listing of the code
knitr::opts_chunk$set(message = FALSE)  ## FALSE suppresses the listing of the message
knitr::opts_chunk$set(warning = FALSE)  ## FALSE suppresses the listing of the warning
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 10)

library(exifr)       ## Read EXIF data from photos
library(flextable)   ## Great table generator
library(gt)          ## Perhaps a more useful table generator
library(dplyr)       ## Data wrangling
library(readr)       ## Read files
library(here)        ## Directory assistance and simplification
library(lubridate)   ## Decode and manipulate dates
library(jpeg)        ## Handle jpeg images
library(magick)      ## Manipulate images
library(ggmap)       ## Wrapper for google maps API
library(patchwork)   ## Assemble and annotate output objects
library(ggplot2)     ## Basic plotting
library(cowplot)     ## ggdraw
library(png)         ## Read PNG files
library(ggrepel)     ## Automatic adjustment of labels
library(sp)          ## Used to calculate the center of a set of points
library(raster)      ## Used to calculate the center of a set of points
library(grobblR)     ## To composite the photos on one page; also convert png to grob
library(numbers)     ## mod function
library(pdftools)    ## Build PDF docs
library(tibble)      ## Used for the add_column function
library(webshot)     ## Needed to help save images
library(ptinpoly)   ## Points inside or outside a polygon

library(BiocManager)
##BiocManager::install("EBImage")
library(EBImage)

## Initialize Google Map key; the is stored in a Project directory.
My_Key <- read_file("P://Hot/Workflow/Workflow/keys/Google_Maps_API_Key.txt")

## Register the Google Maps API Key.
register_google(key = My_Key, account_type = "standard")
```

## Custom Functions

There are several functions that make the processing of photos appear simpler. Several of these are quite general as they are very useful when drawing any kind of map.

* LineWrapper: Inserts "\n" into strings to break them when printing

* tmt_text: Adds text blocks to a map (used inside ggmap)

* tmt_points: Adds point symbols to a map (used inside ggmap)

* tmt_labels: Adds labels to a map (used inside ggmap)

* map_center: Calculate the center of a set of geographic points

* map_zoom: Determine the zoom level for a Google map that includes all the points

```{r functions}

##### FUNCTION LINEWRAPPER #######################################

LineWrapper <- function(x,length=25){
  
## Wraps text with a default of 25 characters
  
## Handle the column that needs to be wrapped
LineWrapper = sapply(lapply(x, strwrap, width=length), paste, collapse=" \n")  
  
} ## end function Linewrapper

##### FUNCTION tmt_text ##########################################

tmt_text <- function(text,texts_default=text_default){
  
  ## Add labels to a map (used inside ggmap)

  ## Fill in any missing columns with default values 
  ## Thanks to Chris Umphlett on StackOverflow (5/3/2019)
  text <- text %>% 
  add_column(!!!texts_default[!names(texts_default) %in% names(.)])
  
  text$text <- str_replace(text$text,"%%","\n")

  ## Generate the labels
  geom_text(data = text,
            aes(x = lon, y = lat, 
                label = text),
                size          = text$size, 
                color         = text$color, 
                fontface      = text$fontface,
                angle         = text$angle) 

  } ## End function tmt_text

##### FUNCTION tmt_points ########################################

tmt_points <- function(points, points_default=point_default){
  
  ## Add points to a map (used inside ggmap)

  ## Fill in any missing columns with default values 
  ## Thanks to Chris Umphlett on StackOverflow (5/3/2019)
  points <- points %>% 
  add_column(!!!points_default[!names(points_default) %in% names(.)])

  ## Note the following convention for symbol shapes.
  ## You can use circle, square, diamond, triangle,& wedge.
  ## These are mapped to pch 21-25.
  
  points$shape <- sub("circle",  "21",points$shape)
  points$shape <- sub("square",  "22",points$shape)
  points$shape <- sub("diamond", "23",points$shape)
  points$shape <- sub("triangle","24",points$shape)
  points$shape <- sub("wedge",   "25",points$shape)
  
  points$shape <- as.numeric(points$shape)

  ## Generate the labels
  geom_point(data = points,
             aes(x = lon, y = lat),
                  size          = points$size, 
                  color         = points$color,              
                  alpha         = points$alpha,                 
                  fill          = points$fill,
                  shape         = points$shape,
                  stroke        = points$stroke)
  
} ## End Function tmt_points

##### FUNCTION tmt_labels #######################################

tmt_labels <- function(labels,labels_default=label_default){
  
  ## Add labels to a map (used inside ggmap)

  ## Fill in any missing columns with default values 
  ## Thanks to Chris Umphlett on StackOverflow (5/3/2019)
  labels <- labels %>% 
  add_column(!!!labels_default[!names(labels_default) %in% names(.)])

  ## Generate the labels
  geom_label_repel(data = labels,
             aes(x = lon, y = lat, 
                 label = text),
                  size          = labels$size, 
                  color         = labels$color,              
                  alpha         = labels$alpha,                 
                  label.r       = labels$label.r,
                  label.size    = labels$label.size,
                  fill          = labels$fill,
                  box.padding   = 0.8,
                  #point.padding = labels$point.padding,
                  segment.color = labels$segment.color,
                  segment.size  = labels$segment.size,
                  min.segment.length = 0.01, 
                  seed          = labels$seed) 

  } ## End function tmt_labels

##### FUNCTION MAP_CENTER #########################################

map_center <- function(data_points){

## Finds the lat,lon of the center of a set of data points (each a lat,lon).  
    
## Required libraries (sp, raster)

## Data.frame data_points must have variables "lat" and "lon". 

## Transfer the data into a new data.frame
dat <- data.frame(lat=data_points$lat, lon=data_points$lon) 

## Put the coordinates into a SpatialPointDataFrame
coordinates(dat) <- ~lon+lat
proj4string(dat) <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
    
## Get the center point coordinates    
cpt <- coordinates(as(extent(dat), "SpatialPolygons"))

cpt <- as.data.frame(cpt) %>% 
  rename(lat = V2, lon = V1)

cpt <- as.list(cpt)

return(cpt)
}  ## end map_center

##### FUNCTION MAP_ZOOM ###########################################

map_zoom <- function(point_data, center_point){
  
## The data.frame data_points must have variables "lat" and "lon".
## Similarly, center_point must also have a "lon" and "lat".
  
## Set starting conditions.
zoom <- 17
outside <- FALSE

## Look at successively zoomed out levels until all points are inside
for (level in seq(from=zoom, to=5, by=-1)){
  
    ## Get a Google map at a particular zoom level
    t_map <- get_googlemap(center=center_point,
                           zoom=level)
    
    ## Parse out the latitude and longitude data of the bounding box
    box <- attr(t_map,"bb")

    ## Rename the bounding box coordinates
    box <- box %>% 
           rename(left = ll.lon, right = ur.lon, bottom = ll.lat, top = ur.lat)
    
    ## Get the coordinates of the vertices of the box
    box_vertices <- c(box$left,box$bottom,
                      box$left,box$top,
                      box$right,box$top,
                      box$right,box$bottom)
    
    ## Make the vertices into matrix values
    b_v_matrix <- matrix(box_vertices,nrow=4,ncol=2,byrow=TRUE)
    d_p_matrix <- matrix(c(point_data$lon,point_data$lat),
                         nrow=nrow(point_data),
                         ncol=2,
                         byrow=FALSE)
    
    ## Using this function looks like overkill, but it is a general solution to the
    ## problem of finding points inside a polygon. It did not seem to work with negative
    ## values for the longitude. So absolute values are used.
    ## res is a vector of all of the points indicating if they are inside or outside.
    res <- pip2d(abs(b_v_matrix),abs(d_p_matrix))
    
    ## Test if any of the points are outside the bounding box (-1 = outside; 1 = inside)
    if (any(res < 0)){outside <- TRUE}
  
    ## The most zoomed-in level is chosen.
    if(outside != TRUE) {
      level <- level - 1
      return(level)
    
      } ## end if outside
    
  } ## end Level loop

} ## end function map_zoom



```

## Photo Information

The overall goal is to process all the photos in a folder and produce a PDF document that summarizes the photo information.




```{r photos}

## The one thing you need to specify
folder   <- "Photos/Tidal_Basin"

#######################################################################

## Make a blank image (used later when images are printed)
blank <- image_blank(width=640, height=480, color = "white", pseudo_image = "", defines = NULL)
image_write(image=blank,path="temp/z_blank.png")

## Read all the files in a folder
files <- list.files(path=here(folder), 
                    pattern="*.jpg", 
                    full.names=TRUE, 
                    recursive=FALSE)

## Count the photos in the folder
no_files <- length(files)

## Prepare a place to store the photo data
out_info <- setNames(data.frame(matrix(ncol = 4, nrow = 0)), 
                     c("number", "location", "lat", "lon"))

table_info <- setNames(data.frame(matrix(ncol = 5, nrow = 0)), 
                     c("number", "file", "location", "date", "time"))

## Process each of the photos, one at a time
for (i in 1:no_files){
 
  ## Here is where the heavy lifting is done
  sheet  <- image_read(files[i])
  
  tags <- c("GPSLatitude","GPSLongitude","DateTimeOriginal")
  
  photo_exif <- read_exif(path=files[i],tags=tags,recursive = FALSE)

  ## Make a small, reference image
  small_photo <- image_scale(sheet,"10%")

  ## Get the location and put in easy-to-use variables
  in_lat <- photo_exif$GPSLatitude
  in_lon <- photo_exif$GPSLongitude
  in_location <- c(in_lon, in_lat)
  
  ## Process the date and time

  in_date <- as.Date(substring(photo_exif$DateTimeOriginal,1,10),"%Y:%m:%d")

  lub_str   <- ymd_hms(photo_exif$DateTimeOriginal) 
  in_hour   <- hour(lub_str)
  in_minute <- minute(lub_str)
  in_minute <- sprintf("%02d",in_minute)
  in_time   <- paste0(in_hour,":",in_minute)
  
  in_date <- as.Date(substring(photo_exif$DateTimeOriginal,1,10),"%Y:%m:%d")
  in_when <- paste0(in_time," on ",weekdays(in_date),", ",
                    format(in_date, format="%B %d, %Y")) 

  ## Get the name of the place
  in_place_name <- revgeocode(in_location)
  in_place_name <- LineWrapper(in_place_name, length = 50)
  
  ## Make a caption with the place name and when the photo was taken
  place_caption <- paste0("Photo: ",i," -- ", in_when,"\n",in_place_name[1])

  ## Composite the image and the caption
  out_img <- image_ggplot(small_photo)
  out_img <- out_img + ggtitle(place_caption)
  
  ## Write the file for later use
  out_name <- paste0("out_",i,".png")
  ggsave(out_name,out_img)
  
  ## Save the basic data about each image for later processing
  out_list <- data.frame(number=i, location=in_place_name[1], lat=in_lat, lon=in_lon)
  out_info <- rbind(out_info, out_list)
  
  ## Build a table to add to the document
  f_name <- basename(files[i])
  table_list <- data.frame(number=i, file = f_name, 
                           location=in_place_name[1], date=in_date, time=in_time)
  table_info <- rbind(table_info, table_list)

 } ## End photo loop

## Create a basic table of the photo information
table_1 <- gt(table_info) %>% 
  tab_header(
    title = paste0("Photo Index: ", folder)
    ) %>% 
  tab_source_note(
    source_note = paste0("Report Prepared: ",Sys.Date())
  ) %>%
  tab_style(
    style = cell_text(v_align="top"),
    locations = cells_body()
    ) %>% 
  opt_row_striping(row_striping = TRUE)

## Put the table as a temporary PNG file
gtsave("loc_table_temp.png",data=table_1,path="temp")

## Use the grob functions to get a PDF version of the table
loc_table <- grob_layout(
  grob_row(
    border = FALSE,
    grob_col(
      border = FALSE,
      aes_list = ga_list(
        maintain_aspect_ratio = TRUE
        ),
      "temp/loc_table_temp.png"
      )
    ),
  height = 279.4,
  width = 215.9
  ) 

## Put the grob out as a PDF file
grob_to_pdf(
  list(loc_table),
  file_name = file.path("temp", "location_table.pdf"),
  meta_data_title = "Grob approach to PDFs"
  )

## Clean up by removing the temporary PNG file
file.remove("temp/loc_table_temp.png")
```

## Make a Montage

This puts six images on a page (as a PDF file).

The routine needs to be generalized so that any number of images will be made into 6-up pages.



```{r}
## Required libraries (grobblR, numbers)

## Round the number of files up to a multiple of 6
files_ceiling <- ceiling(7/6) * 6

## Process each labeled small picture, one at a time
page_no <- 0

for(i in 1:files_ceiling){

## Up to six images per page
  ## Count by six
  img_no <- mod(i-1,6) + 1
 
  ## Build a name for the temporary grob image
  img_name <- paste0("out_",i,".png")
  if(!file.exists(img_name))img_name <- "z_blank.png"

  if(img_no == 1) img1g <- grob_image(img_name)
  if(img_no == 2) img2g <- grob_image(img_name)
  if(img_no == 3) img3g <- grob_image(img_name)
  if(img_no == 4) img4g <- grob_image(img_name)
  if(img_no == 5) img5g <- grob_image(img_name)
  if(img_no == 6) img6g <- grob_image(img_name)
  
  ## Things to do when the images fill a page
  if(img_no == 6){
    
    ## Keep track of the page count
    page_no <- page_no + 1
    
    ## Make a montage
    page <- grob_layout(
      grob_row(
        grob_col(img1g),
        grob_col(img2g)
        ),
      grob_row(
        grob_col(img3g),
        grob_col(img4g)
        ),
      grob_row(
        grob_col(img5g),
        grob_col(img6g)
        ),
      height = 279.4, 
      width = 215.9)   

    ## File name for the PDF file
    page_name <- paste0("temp/Photo_map",page_no,".pdf")

    ## Put the page into a PDF file
    grob_to_pdf(
      list(page),
      file_name = page_name,
      meta_data_title = paste0("Photo Map Test PDF page ",page_no))

    } ## end img_no section

  } ## end for loop

  

```

## Some Defaults


```{r defaults}

## Label Defaults
## Note: "lines" is a unit of measure for some dimensions (default)
label_default       <- NULL          ## Needed to get started
label_default$size  <- 4             ## text size        
label_default$color <- "black"       ## text color
label_default$alpha <- 0.9           ## transparency
label_default$label.r <- unit(0.3, "lines") ## rounded corner radius 
label_default$label.size    <- 0.5   ## label border size (mm)
label_default$fill    <- "white"     ## label background color
#label_default$box_padding <- 0.8     ## from point location
#label_default$point.padding <- 0     ## space around point
label_default$segment.color <- "red" ## line from point to label
label_default$segment.size  <- 1
label_default$seed          <- 1002  ## randomizes without a value 

## Point Defaults
point_default <- NULL                ## Needed to get started
point_default$size   <- 3            ## symbols size
point_default$color  <- "black"      ## outside border
point_default$alpha  <- 0.9          ## point transparency
point_default$fill   <- "red"        ## inside color
point_default$shape  <- "circle"     ## circle (symbol 21)
point_default$stroke <- 1            ## border thickness

## Text default
text_default          <- NULL       ## Needed to get started
text_default$size     <- 4          ## text size
text_default$color    <- "black"    ## text color
text_default$fontface <- "plain"    ## alt: italics, bold
text_default$angle    <- 0          ## negative is clockwise

## Google Maps style defaults
## Use these to hide map layers
no_poi        <- "feature:poi|element:labels|visibility:off"
no_transit    <- "feature:transit|element:labels|visibility:off"
no_water      <- "feature:water|element:labels|visibility:off"
no_road       <- "feature:road|element:labels|visibility:off"
no_landscape  <- "feature:landscape|element:labels|visibility:off"
no_administrative <- "feature:administrative|element:labels|visibility:off"
no_country    <- "feature:administrative.country|element:labels|visibility:off"
hide_all      <- "feature:all|element:labels|visibility:off"

```

## Photo Locations

This retrieves a Google Map and places the photo number in the proper location as a map overlay.


```{r refmap}

## Use the data points to make a reference map

## Find the center of the data points
cpt <- map_center(out_info)

cpt <- c(cpt$lon, cpt$lat)

## Calculate the best zoom level to hold all of the points
best_zoom <- map_zoom(point_data = out_info, center_point = cpt)

## Use the map center (cpt) and the data points (out_info) to make the map

## Rename a column in the data for use with the mapping functions.
point_data <- out_info %>% 
  rename(text = number)

## Retrieve a satellite map.
temp_map <- get_googlemap(center= cpt,
                        zoom= best_zoom,
                        style=no_country, maptype = "satellite")

## Display the map.
## ggmap(temp_map)

## Plot the labels on the map with ggmap using the custom functions. 
final_map <- ggmap(temp_map) +
   tmt_points(points=point_data,
              points_default=point_default) +
   tmt_labels(point_data) +
   labs(x="",y="") +
   theme(panel.border=element_rect(color="Black", fill=NA, size=2))

final_map

ggsave(filename = "temp/report_map.pdf",final_map,width=8,height=10)

```

## Assemble the PDF

PDF files of each of the pages have been stored in the `temp` folder. 

The file names were chosen so that the pages will be combined in the proper order. (Alphabetical)

The assembled PDF is saved in the folder with the original images.

```{r assemble}

pdf_files <- list.files(path="temp", 
                    pattern="*.pdf", 
                    full.names=TRUE, 
                    recursive=FALSE)

report <- paste0(folder,"/photo_report.pdf")

pdf_combine(pdf_files, report)

```

