
# Choropleth Maps

Choropleth maps show themes based on predefined regions (e.g., countries, states, counties). Colors, or shades of a single color, are used to differentiate areas. The color of a region corresponds to a data characteristic or range of data values.

## Initialize the Libraries and Data

There are a few things needed to setup the environment.

The examples will make use of the boundaries for the Hawaiian Islands. These have been stored in a (local) folder. The outlines are retrieved here.

```{r}
#| label:   core-choro
#| message: false
#| echo:    true
#| warning: false

## Set an option for the read_csv function
options(readr.show_col_types = FALSE) ## Suppress warning msg

## Initialize the Standard Libraries
library(readr)
library(ggplot2)
library(dplyr)
library(gt)

## Install the sitemaps package (uncomment; run once)
## library(devtools)
## install_github("kimbridges/sitemaps")
## install_github("kimbridges/plainmaps")

## Sitemaps
library(sitemaps)
library(plainmaps)

## Map Libraries
library(mapdata)
library(sf)

## Initialize defaults for Sitemaps
column <- site_styles()
hide   <- site_google_hides()

## Themes
plain_background_theme <- theme(
  panel.border     = element_blank(), 
  panel.grid       = element_blank(),
  panel.background = element_blank(),
  axis.text        = element_blank(),
  axis.ticks       = element_blank(),
  axis.title       = element_blank()) 

island_map_theme <- theme(
  panel.border = element_rect(color     = "black", 
                              fill      = NA, 
                              linewidth = 0.7),   
  panel.grid     = element_blank(),
  panel.background = element_rect(fill = "lightsteelblue1"),
  axis.text      = element_blank(), 
  axis.ticks     = element_blank(), 
  axis.title     = element_blank())


## Read stored border data
oahu_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/oahu_outline.csv")

hawaii_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/hawaii_outline.csv")

maui_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/maui_outline.csv")

kahoolawe_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/kahoolawe_outline.csv")

lanai_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/lanai_outline.csv")

molokai_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/molokai_outline.csv")

kauai_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/kauai_outline.csv")

niihau_simple <- read_csv(col_names = TRUE, 
                    show_col_types = FALSE,
                    file = "hawaii_400/niihau_outline.csv")

```

## Choropleth Maps for Hawai\`i

We can demonstrate a typical choropleth map using the political classification of the Hawaiian Islands.

Each island is a "predefined region." The color code represents the county for each island.

```{r}
#| label:   islands
#| message: false
#| echo:    true
#| warning: false

## Assign islands to counties by using color.
fill <- NULL
fill$hawaii <- "darkseagreen3"
fill$maui   <- "goldenrod1" 
fill$oahu   <- "peachpuff3"
fill$kauai  <- "lightsalmon1"

## Plot each of the islands with its fill color.
ggplot() + 
  geom_polygon(data=hawaii_simple, aes(x=lon,y=lat),
               fill=fill$hawaii, 
               linewidth=0.4,
               color="black") +
  geom_polygon(data=oahu_simple, aes(x=lon,y=lat),
               fill=fill$oahu, 
               linewidth=0.4,
               color="black") +
  geom_polygon(data=lanai_simple, aes(x=lon,y=lat),
               fill=fill$maui, 
               linewidth=0.4,
               color="black") +
  geom_polygon(data=kauai_simple, aes(x=lon,y=lat),
               fill=fill$kauai, 
               linewidth=0.4,
               color="black") +
  geom_polygon(data=niihau_simple, aes(x=lon,y=lat),
               fill=fill$kauai, 
               linewidth=0.4,
               color="black") +
  geom_polygon(data=kahoolawe_simple, aes(x=lon,y=lat),
               fill=fill$maui, 
               linewidth=0.4,
               color="black") +
  geom_polygon(data=maui_simple, aes(x=lon,y=lat),
               fill=fill$maui, 
               linewidth=0.4,
               color="black") +
  geom_polygon(data=molokai_simple, aes(x=lon,y=lat),
               fill=fill$maui, 
               linewidth=0.4,
               color="black") +
  coord_fixed(1.1) +
  island_map_theme
  
```

## Using Data to Assign Colors

Note that in this example, the colors are not assigned directly in the `ggplot` step. Instead, data values point to specific items in a color table.

This approach is especially valuable when the `color_index` values are computed. Here, these values need to be integers that correspond to the range of the items in the color table.

This turns the procedure into a more useful tool for displaying choropleth maps. Note the following characteristics:

-   The color are given with an index value (called `quartile` here). It is straightforward to change the colors in just this one place.

-   There are four `quartile` values. This is because a quantitative value (`density`) is being used as the basis of the colors. The division is into quartiles (four values based on an order list divided into 25% groups).

-   The population density is calculated.

-   The code for the data table, used for data confirmation, looks very long. However, there are formatting elements shown here for things like numeric formats and footnotes. It is a good habit to make complete (i.e., well annotated) data tables. Usually, it's just a matter of copying the formatting code used in other tables.

-   A new variable (`quartile`) is calculate using Sitemap's `site_cuts` function. You can get more information on this function in the Site Maps document.

-   The color for each island is merged into the basic data table by matching the common values for `quartile`.

-   The code used in the `ggplot` function is simplified by using the `update_geom_defaults` function to set values that will be used in each of the `geom_polygon` functions.

-   The `color` for each island's fill value is retrieved by indexing rows the the island `name`.

-   The legend is added with a `scale_fill_manual` function. The values for this are initialized in the Data and Parameter Input section.

All of these steps make a generalizable procedure that can be used for other chloropleth plots.

The code is divided into large sections to help identify the overall visualization strategy.

```{r}
#| label:   colorcode
#| message: false
#| echo:    true
#| warning: false

## Data and Parameter Input ################################

## Color table: List of colors and their index values (quartile).
color_table <- 
  read_csv(file =
    "color,   quartile
     beige,   1 
     wheat2,  2
     wheat3,  3         
     wheat4,  4")

## Assign values for the legend.
ct_value <- color_table$quartile
ct_color <- color_table$color
legend_title <- "Population\nDensity\nQuartiles"

## Read the names and data values for the islands.
island <- read_csv(file =
  "name  , population, area
   Oahu,   953000,      597
   Hawaii, 185000,     4208
   Maui,   155000,     1173
   Kauai,   67000,      620
   Molokai,  7300,      260
   Lanai,    3100,      141
   Niihau,    130,       72
   Kahoolawe,   0,       45")

## Note: Island boundaries were set in the Initialization chunk.

## Data Manipulation #######################################

## Calculate the population density for each island
island <- island |>
  mutate(density = population/area) 

## Calculate an index for the population density
## This uses the Sitemaps site_cuts function to calculate quartiles.
island$quartile <- site_cuts(quant_var = island$density,
     cuttype = "quartiles4")

## Merge the colors using the quartile values as an index
island <- merge(island, color_table, by="quartile")

## Data Check #############################################

## Show a well-formatted data table
gt(island) |>
  ## Number formats
  fmt_number(
    columns  = c(population, area),  
    use_seps = TRUE,
    decimals = 0) |> 
  fmt_number(
    columns = density,
    use_seps = TRUE,
    decimals = 2) |>
  ## Footnotes
  tab_footnote(
    footnote = "miles\U00B2", ## Symbol for squared
    locations = cells_column_labels(columns=area)) |> 
  tab_footnote(
    footnote = "people/mile\U00B2",
    locations = cells_column_labels(columns=density)) |> 
  ## Data source
  tab_source_note(source_note="Source: Gemini AI query")

## Plot Section ################################################
 
## Make a set of default values for the geom_polygon parameters.
update_geom_defaults("polygon", 
                     list(color = "black", linewidth = 0.4))

## Plot each of the islands using the color as fill.
island_map <- ggplot() + 
  geom_polygon(data=hawaii_simple, aes(x=lon,y=lat,
      fill=island$color[island$name=="Hawaii"])) +
  geom_polygon(data=maui_simple, aes(x=lon,y=lat, 
      fill=island$color[island$name=="Maui"])) +
  geom_polygon(data=molokai_simple, aes(x=lon,y=lat,
      fill=island$color[island$name=="Molokai"])) +
  geom_polygon(data=kahoolawe_simple, aes(x=lon,y=lat, 
      fill=island$color[island$name=="Kahoolawe"]))+
  geom_polygon(data=lanai_simple, aes(x=lon,y=lat, 
      fill=island$color[island$name=="Lanai"])) +
  geom_polygon(data=oahu_simple, aes(x=lon,y=lat,
      fill=island$color[island$name=="Oahu"])) +
  geom_polygon(data=kauai_simple, aes(x=lon,y=lat,
      fill=island$color[island$name=="Kauai"])) + 
  geom_polygon(data=niihau_simple, aes(x=lon,y=lat,
      fill=island$color[island$name=="Niihau"])) + 
  coord_fixed(1.1) +
  island_map_theme

## Create the map with a legend
choro_map <- island_map +
  scale_fill_manual(legend_title,
                    labels = ct_value,
                    values = ct_color)

## Plot the map and legend
choro_map

```

## Annotation

The functions in `Sitemaps` can be used to add text annotations to the choropleth map.

Text position coordinates can be copied from Google Maps by right-clicking on a location and copying the coordinates from the pop-up box by left-clicking on the coordinates. You get a message that the coordinates have been copied to the clipboard.

Recall that the `text` is centered on the `lat, lon` coordinates unless you change the `justify` value to `left` or `right`. **\[NOTE: JUSTIFY DOESN'T SEEM TO BE WORKING RIGHT IN SITEMAPS\]**

The example below uses the Population Density map from the previous chunk.

```{r}
#| label:   annotate
#| message: false
#| echo:    true
#| warning: false

## Create a file of names for the annotation.
annotation <- read_csv(file =
      "text,   lat,       lon,         fontface, name_text_size
  Ni`ihau,     21.67626, -160.09480,   plain,    4
  Kaua`i,      22.11367, -158.96844,   plain,    4
  O`ahu,       21.35112, -158.41976,   plain,    4
  Moloka`i,    21.35607, -156.68232,   plain,    4
  Lāna`i,      20.78355, -157.30721,   plain,    4
  Maui,        20.97766, -155.93282,   plain,    4
  Kaho`olawe,  20.40621, -157.01449,   plain,    4
  Hawai`i,     19.55664, -156.38691,   plain,    4
  Pacific Ocean, 19.74548, -158.75696, italic,   5")

## Plot the map with the annotation.
choro_map +
  site_names(datatable = annotation)

```
