---
title: "Sandbox"
editor: visual
---

rgeoboundaries

```{r}
library(rnaturalearth)
library(rnaturalearthdata)
library(tidyverse)
library(sf)
library(jsonlite)

test <- ne_countries(scale= 10, 
                     returnclass = "sf") |>
 filter(su_a3 == "WLF")

test_full <- st_geometry(test)

ggplot() + 
  labs(title = "Test: 10 Resolution") +
  geom_sf(data=test_full) 

```

Whittaker Diagram

From Claude 3 


```{r}
# Install and load the required libraries
#install.packages("ggplot2")
library(ggplot2)

# Define the temperature and precipitation ranges
temp_range <- seq(-15, 30, length.out = 100)
precip_range <- seq(0, 400, length.out = 100)

# Create a data frame with all combinations of temperature and precipitation
data <- expand.grid(temp = temp_range, precip = precip_range)

# Define the biome labels and their corresponding regions
biomes <- list(
  list("Tundra", data$temp < -5),
  list("Boreal Forest", data$temp >= -5 & data$temp < 5 & data$precip > 50),
  list("Temperate Rainforest", data$temp >= 5 & data$temp < 15 & data$precip > 200),
  list("Temperate Seasonal Forest", data$temp >= 5 & data$temp < 15 & data$precip > 50 & data$precip <= 200),
  list("Temperate Grassland/Desert", data$temp >= 5 & data$temp < 15 & data$precip <= 50),
  list("Subtropical Desert", data$temp >= 15 & data$precip < 25),
  list("Tropical Rainforest", data$temp >= 20 & data$precip > 200),
  list("Tropical Seasonal Forest/Savanna", data$temp >= 20 & data$precip > 50 & data$precip <= 200),
  list("Subtropical Rainforest", data$temp >= 15 & data$temp < 20 & data$precip > 200)
)

# Assign biome labels to the data frame
data$biome <- "Other"
for (biome in biomes) {
  data$biome[biome[[2]]] <- biome[[1]]
}

# Create the Whittaker diagram using ggplot2
diagram <- ggplot(data, aes(x = precip, y = temp, fill = biome)) +
  geom_raster(alpha = 0.7) +
  scale_fill_manual(values = c(
    "Tundra" = "#B2B2FF",
    "Boreal Forest" = "#8080FF",
    "Temperate Rainforest" = "#4D4DFF",
    "Temperate Seasonal Forest" = "#1A1AFF",
    "Temperate Grassland/Desert" = "#E6E6E6",
    "Subtropical Desert" = "#FFFF4D",
    "Tropical Rainforest" = "#00994D",
    "Tropical Seasonal Forest/Savanna" = "#80FF80",
    "Subtropical Rainforest" = "#00CC66",
    "Other" = "white"
  )) +
  labs(
    title = "Whittaker Biome Diagram",
    x = "Annual Precipitation (cm)",
    y = "Average Annual Temperature (°C)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Display the diagram
print(diagram)
```

This R script follows a similar approach to the Python program:

1. We load the `ggplot2` library, which is used for creating the diagram.

2. We define the temperature and precipitation ranges using `seq()` to create evenly spaced values.

3. We create a data frame called `data` using `expand.grid()` to generate all combinations of temperature and precipitation values.

4. We define a list of biomes, where each biome is represented by a list containing its label and the corresponding condition based on temperature and precipitation.

5. We assign biome labels to the `data` data frame by iterating over the biomes and updating the `biome` column based on the conditions.

6. We create the Whittaker diagram using `ggplot()`:
   - We specify the `data` data frame and map the `precip` and `temp` columns to the x and y aesthetics, respectively.
   - We use `geom_raster()` to create the raster plot and set the `alpha` parameter to control transparency.
   - We use `scale_fill_manual()` to assign specific colors to each biome.
   - We add labels for the title, x-axis, and y-axis using `labs()`.
   - We customize the theme using `theme_minimal()` and set the legend position to the bottom.

7. Finally, we display the diagram using `print()`.

When you run this R script, it will generate a Whittaker diagram showing the distribution of biomes based on temperature and precipitation, similar to the Python program.


## Periodogram

```{r}

library(TSA)
library(readr)

# Create a sample time series data
ts_data <- read_csv(file="sandbox/KimPuzzle1.csv")

ts_test <- ts_data$data3


ts_test <- na.omit(ts_test)

ts_test

# Perform the periodogram analysis
periodogram(ts_test)


```
WFOR package

https://github.com/rogerhyam/wfor

```{r}
## Set an option for the read_csv function
options(readr.show_col_types = FALSE) 

library(devtools)
##devtools::install_github("rogerhyam/wfor")
library(wfor)
library(readr)
library(gt)

##  WARNING: VERY LONG LIST!!!!!
# auto match the Belgian Magnoliopsida test data
##mags_example <- wfo_match_df_names(Belgian_Magnoliopsida_sp_ssp_var_2011, name_col="scientific_name", authors_col="authorship", interactive=FALSE)

produce <- read_csv(file=
    "Common_Name,Sci_Name,Author
plum,Prunus domestica,L.
lettuce,Lactuca sativa,L.
chard,Beta vulgaris subsp. vulgaris,L.
tomato,Solanum lycopersicum,L.
orange,Citrus x sinensis,(L.) Osbeck
celery,Apium graveolens,L.
dragon fruit,Hylocereus undatus,(Haw.) Britton & Rose
papaya,Carica papaya,L.
langoustine,Nephrops norvegicus,(L.)
raspberry,Rubus idaeus,L.
cherimoya,Annona cherimola,Mill.")

gt(produce)

answer <- wfo_match_df_names(produce, name_col="Sci_Name", authors_col="Author", interactive=FALSE,
                             fallback_to_genus = TRUE)

answer

wfo_stat_matches(answer)

```

O'ahu Botanical Gardens

```{r}

library(sitemaps)
library(ggmap)

data <- read_csv(file=
  "garden
    Ho'omaluhia Botanical Garden
    Waimea Valley
    Koko Crater Botanical Gardens
    Foster Botanical Garden
    Moanalua Gardens
    Lili'uokalani Botanical Garden
    Mānoa Heritage Center
    Halawa Xeriscape Garden
    Wahiawā Botanical Garden
    Lyon Arboretum")

## Geocode the locations.
## Prepare the data.
location <- paste0(data$garden,", Honolulu, Hawaii, USA") 

## Get the coordinates from the location.
coord <- geocode(location, output = "latlon")

## Create an ID column.
ID <- c(1:nrow(data))

## Number & add the coordinates to the original data.
data <- cbind(ID,data,coord)

## Show the data table.
gt(data) %>% 
  fmt_number(columns = c(lat,lon), decimals = 5)



```

