
# Tessellations

BIG PROBLEMS -- DON'T USE UNTIL FIXED

Here is the problem. You have a set of points. Each point belongs to a category. For now, let's assign each point a color. The goal is to extend the color outward from each point until it bumps into a color coming from another point. This makes a mosaic of color. Each colored cell in the mosaic has a data point at its center. This is called a "Voronoi tessellation."

You can see that a tessellation divides a plane into regions based on the distances between a set of points. The visualization of this process produces a "Voronoi diagram."

Creating a Voronoi diagram occurs inside a boundary. This outline may be a square or a polygon.

Voronoi tessellations are an interesting way to fill a boundary with colors based on data values at geographic locations.

## Initializing the Packages

As usual, it is necessary to initialize some R packages to provide the functions needed to handle the data and create the visualizations.

```{r}
#| label:   core-tess
#| message: false
#| echo:    true
#| warning: false


## Set an option for the read_csv function
options(readr.show_col_types = FALSE) ## Suppress warning msg

## Initialize the standard packages
library(readr)
library(ggplot2)
library(dplyr)
library(gt)

## Sitemaps
library(sitemaps)
library(plainmaps)

## Useful tools
library(remotes)

## Initialize specialized packages
## remotes::install_github("garretrc/ggvoronoi", dependencies = TRUE, build_opts = c("--no-resave-data"))
library(ggvoronoi)

library(ggforce)

## Initialize for Sitemaps
column <- site_styles()
hide   <- site_google_hides()

## Read stored border data
oahu_border_400 <- read_csv(file = 
                   "hawaii_400/oahu_outline.csv")

## Themes
plain_background_theme <- theme(
  panel.border     = element_blank(), 
  panel.grid       = element_blank(),
  panel.background = element_blank(),
  axis.text        = element_blank(),
  axis.ticks       = element_blank(),
  axis.title       = element_blank()) 

```

## An Example Tessellation

## Create a Tesselation

Tessellations are built inside a bounding area. For our purposes, this is an area on a map. Inside this area there are data points. Each data point represents a classification unit (such as a forest type, a weather station, or a "magnet location" like a school).

1.  Outline the area.

2.  Define the units and their central locations (centroids).

3.  Create and plot the Voronoi tessellation.

```{r}
#| label:   step1
#| message: false
#| echo:    true
#| warning: false

## STEP 1: Outline the area

## Outline data obtained from Google Maps.
## Google maps often gives order of the coordinates "backwards".
kakaako <- read_csv(file =
  "lat,      lon
   21.29581, -157.86933
   21.29045, -157.86118
   21.29449, -157.85736
   21.29317, -157.85504
   21.29125, -157.85616
   21.29217, -157.85826
   21.29105, -157.85937
   21.29085, -157.85577
   21.29137, -157.85543
   21.28893, -157.84852
   21.29993, -157.84414
   21.30513, -157.85834
   21.30209, -157.86319
   21.30317, -157.86422")

## Swap the order of the columns in the outline.
## The expected order is x, y, group. (Group handles "islands".)
## Make sure the names of the columns are correct.
kakaako <- data.frame(kakaako$lon, kakaako$lat, "A")
names(kakaako) <- c("lon", "lat", "group")

## Check the outline
ggplot(kakaako) +
  geom_polygon(mapping=aes(x=lon, y=lat),fill = "lightgrey", 
               color = "black") +
  geom_point(aes(x=lon, y=lat, color = "red"))
```

```{r}
#| label:   step2
#| message: false
#| echo:    true
#| warning: false

## STEP 2: Define the units and their centroids

## These are the centroid locations of the different areas.
districts <- read_csv(file = 
      "type,          lat,      lon
       shipping,      21.30037, -157.86586
       industrial,    21.29842, -157.85641
       auto sales,    21.29653, -157.86079
       recreation,    21.29319, -157.86435
       recreation,    21.29057, -157.85146
       education,     21.29891, -157.84938
       auto sales,    21.29956, -157.85313
       recreation,    21.29924, -157.85847
       dining,        21.29865, -157.86144
       dining,        21.30169, -157.86259
       entertainment, 21.29432, -157.85372
       industrial,    21.29466, -157.84918
       industrial,    21.29677, -157.85195
       shipping,      21.29253, -157.85973")

## Create a lookup table for the color assignments.
lookup_list <- read_csv(file = 
     "type,           tile_color
      dining,         gold
      education,      green
      entertainment,  lightgreen
      industrial,     blue
      recreation,     purple
      shipping,       magenta
      auto sales,     tomato")

## Merge the look-up table with the districts table.
districts <- merge(districts, lookup_list, by="type")
```

```{r}
#| label:   step3
#| message: false
#| echo:    true
#| warning: false

## STEP 3: Create and plot the voronoi tessellation

## Plot the Voronoi tiles.
ggplot(districts) +
  geom_voronoi(mapping=aes(x=lon,y=lat,fill=type), 
               outline=kakaako, 
               color="black") +
  scale_fill_discrete(name = "District", 
                      labels = unique(districts$type)) +
  scale_color_manual(values=lookup_list$tile_color)

```

## Enhance the Tessellation

Adjacent areas that are the same color don't need to have a boundary drawn between the data location (centroids).

The next example modifies the previous code to eliminate the internal boundaries.

THIS IS NOT WORKING

Alternative

```{r}

colnames(kakaako) <- c("x","y")

kakaako2 <- dplyr::select(kakaako, x, y)
kakaako2 <- data.frame(kakaako2)


ggplot(districts, aes(lon, lat, group = -1L)) +
  geom_voronoi_tile(aes(fill = type), colour = "black",
                    linetype = 0,
                    bound=kakaako2) +
  geom_polygon(data=kakaako2, aes(x,y),
               linetype="solid",
               color="black",fill=NA) +
  plain_background_theme
                    
```


```{r}
#| label:   merge
#| message: false
#| echo:    true
#| warning: false

## This version visually merges the areas that are equal and adjacent.
## The area is outlined with geom_polygon.
## Lots of theme code used to eliminate the background stuff.

vor <- ggplot(districts, aes(x=lon, y=lat, fill=type)) +
  geom_voronoi(outline=kakaako) +
  geom_polygon(data=kakaako,mapping = aes(x=lon, y=lat), 
               color = "black", fill=NA) +
  scale_fill_discrete(name = "District", 
                      labels = unique(districts$type)) +
  scale_color_manual(values=lookup_list$tile_color) +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()) +
  labs(x="",y="")

vor
```

## Market Draw

Colors are not mapping properly.

```{r}
#| label:   markets
#| message: false
#| echo:    true
#| warning: false

## Locations of Costco Warehouse stores on O`ahu.
stores <- read_csv(file = 
 "name,       address,             lat,      lon,        tile_color
  Iwilei,     525 Alakawa St,      21.30893, -157.85833, red
  Hawaii Kai, 333A Keahole St,     21.33105, -157.83748, blue
  Waipahu,    94-1231 Ka Uka Blvd, 21.42206, -158.00944, orange
  Kapolei,    4589 Kapolei Pkwy,   21.34785, -158.02916, yellow")

## Print a data confirmation table.
gt(stores)

## Confirm the map border.
ggplot(oahu_border_400) +
  geom_polygon(mapping=aes(x=lon, y=lat),fill = "lightgrey", 
               color = "black") +
  coord_fixed(1.1)

## Plot the Voronoi tiles.
ggplot(stores, aes(x=lon,y=lat,fill=tile_color)) +
       geom_voronoi(outline=oahu_border_400, 
                    color="black") +
 scale_fill_discrete(name   = "Store", 
                     labels = stores$name)  +
 scale_color_manual(values=stores$tile_color,
                    labels=stores$name) +
 coord_fixed(1.1)

```

Another Try

```{r}
schools <- readr::read_csv(file = 
"number, school,	  lon,	      lat,      color
1, Roosevelt,  −157.83743,	21.31061, red
2, Kaiser,	    −157.69538,	21.28552, blue
3, Kalani,	    −157.77375,	21.27865, green
4, Moanalua,	  −157.90036,	21.34712, yellow
5, Kalaheo,	  −157.75689,	21.40925, purple
6, Mililani,	  −158.01056,	21.45252, orange
7, Pearl City,	−157.97152,	21.39104, brown
8, McKinley,		−157.84824,	21.29900, wheat1
9, Radford,    −157.92850,	21.35984, goldenrod
10, Aiea,       −157.93008,	21.38490, peach
11, Farrington, -157.85056, 21.30472, red
12, Campbell,   -158.00722, 21.32250, blue
13, Kailua,     -157.74056, 21.39167, green
14, Waipahu,    -158.0091,  21.3866,  aqua
15, Waianae,    -158.1858,  21.4458,  pink
16, Waialua,    -158.1283,  21.5732,  scarlet
17, Nanakuli,   -158.1545,  21.3900,  coral
18, Leilehua,   -158.0203,  21.4506,  darkseagreen1
19, Kapolei,    -158.0618,  21.3353,  azure2
20, Castle,     -157.7394,  21.4083,  bisque4
21, Kaimuki,    -157.8021,  21.2905,  darkgrey")

schools <- dplyr::arrange(schools, school)

gt(schools)

## Boundary
## Make the labels x, y
outline <- readr::read_csv("inst/extdata/oahu_outline.csv")
colnames(outline) <- c("x", "y" )

ggplot(schools, aes(x=lon, y=lat, group = -1L)) +
  geom_voronoi_segment(colour = "black") + 
  ##geom_voronoi_tile(aes(fill=number), color="black") +
  ##                  bound=outline) +
 geom_point() 
 ## coord_fixed(1.1) 
```

