# Enhancing Braun–Blanquet Tables

## Color, Site Metadata and Symbolism

### Using Color as a Presentation Aid

Color can be a useful aid when inspecting and presenting Braun–Blanquet tables, but it should be applied carefully. In this workflow, color is treated strictly as a **presentation enhancement**. It does not alter the data, the ordering, or the structural metrics, and it is not required for interpretation.

The primary role of color here is to support the eye in detecting patterns that are already present in the table. Light shading helps distinguish low from high cover or abundance values, especially in larger tables where visual scanning becomes more demanding. At the same time, the table must remain readable without color; group boundaries, labels, and numerical values should still carry the main interpretive weight.

For this reason, the examples in this chapter use a **subtle, monotonic color ramp** applied after ordering and annotation. Zeros are displayed as empty cells, and color is applied only to nonzero values. This preserves the familiar appearance of Braun–Blanquet tables while providing additional visual guidance.

The color scale is defined using a **fixed domain** corresponding to the cover or abundance scale used in the data. This is an intentional choice. Fixing the domain ensures that color has a consistent meaning across tables and avoids exaggerating differences when a particular table happens to contain only low values. Users working with different scales can adjust the domain explicitly as needed.

To keep each chapter self-contained, the table shown here is a simplified example constructed locally. The same approach applies to more complex tables and to tables produced in other chapters.

### Adding color

Color can be added as an optional overlay using standard `gt` functionality. The example below uses a fixed color domain corresponding to the Braun–Blanquet cover/abundance scale. This preserves interpretability and comparability across tables.

#### Creating an uncolored table

This uses a simplified workflow that doesn't involve grouping or other enhancements.

```{r}

library(readr)
library(gt)
library(bbstruct)
library(magick)

options(readr.show_col_types = FALSE)

## Read the raw data 
raw_data <- readr::read_csv(file =
"species, S01, S02, S03, S04, S05, S06, S07, S08
 sp_A,    3,   3,   2,   1,   0,   0,   0,   0
 sp_B,    2,   3,   1,   0,   0,   0,   0,   0
 sp_C,    2,   1,   2,   3,   1,   0,   0,   0
 sp_D,    0,   1,   1,   3,   3,   0,   0,   0
 sp_E,    0,   0,   1,   3,   2,   0,   0,   0
 sp_F,    0,   0,   1,   0,   1,   0,   3,   0
 sp_G,    0,   0,   0,   0,   0,   1,   0,   0
 sp_H,    0,   0,   0,   0,   0,   1,   0,   3"
)

## Inspect raw data
gt(raw_data)

## Convert to Braun-Blanquet table format
bb_table <- bb_from_raw_data(raw_data)

## Reorder the table (baseline view)
bb_ordered <- bb_reorder(bb_table, method = "blind")

## Get the table for the ordered BB data
gt_tbl <- bb_gt(bb_ordered, 
                title = "Ordered table", 
                blank_zeros = TRUE)

## Show the table
gt_tbl

```

#### Adding a color overlay

This overlay builds on the table just created.

The colorizing function, bb_gt_color, lets you control the number of steps in the color ramp. The max_val should be equal to or greater than the largest value in the cell data.

```{r}

## Add a color overlay to the table based on cover/abundance
gt_tbl2 <- gt_tbl |> 
  bb_gt_color(max_val = 3)

## Show the table
gt_tbl2

```

Having a value that is larger than the data maximum spreads out the colors so that they are more subtle. You can also change the color of the ramp.

```{r}

## Add a color overlay to the table based on cover/abundance
gt_tbl2 <- gt_tbl |> 
  bb_gt_color(max_val = 7, high = "orange")

## Show the table
gt_tbl2


```

#### Saving the table

This is an optional step. There are two versions. The first creates a color PNG file. The parameters make this a higher resolution file.

```{r}

## Save the table as png file
gtsave(gt_tbl2,filename = "color_table.png", 
       expand = 10,
       vwidth = 1600,
       vheight = 1000,
       zoom = 2)

```

The second example creates a color PDF file by converting the png file created in the previous step. This is complex due to the failure of the pdf file saving routine to sometimes skip the color. Converting the file this way works.

```{r}

## Convert the png table to a pdf file (using package magick)
image_read("color_table.png") |>
  image_write("color_table.pdf", format = "pdf")


```

### Site Metadata

Site metadata are recorded in a separate table with one row per site. This table provides environmental and contextual attributes used for interpretation. In later steps, a single metadata attribute (e.g., elevation) can be added as an annotated row above the species data in the BB table.

#### Building an example site metadata table

Just as with the species data, you can provide the site metadata in several ways. In the spirit of showing details in a standard way, we'll build a simple table here as an example.

```{r}

library(readr)
library(gt)
library(bbstruct)

## Read the site metadata
site_metadata <- readr::read_csv(file = 
    "name,            id,  elevation, slope, aspect,  soil
     Airport Lot,     S01,  123,        5°,     NE,     U
     Hayes Road,      S02,  231,       12°,     N,      U
     Johnson Ranch,   S03,  485,       18°,     SE,     M
     Hwy 76 Reststop, S04,  729,        4°,     S,      M
     Baker Trailhead, S05, 1303,       19°,     SE,     I
     Cliffside Camp,  S06, 1545,       21°,     E,      I
     Heart Lake,      S07, 1840,       17°,     S,      I
     Alpine Meadow,   S08, 2135,        9°,     N,      I")

## Make an inspection table
meta_tbl <- gt(site_metadata) |>
  tab_header(title="Raw Site Metadata Table") |>
  tab_footnote(
    footnote = "meters",
    locations = cells_column_labels("elevation"))  |>
  tab_footnote(
    footnote = "degrees",
    locations = cells_column_labels("slope"))  |>
  tab_footnote(
    footnote = "key: U=Ultisol; M=Mollisol; I=Inceptisol",
    locations = cells_column_labels("soil"))

## Show site metadata table
meta_tbl

## Save the table as a pdf file (optional)
gtsave(meta_tbl,filename = "site_metadata.pdf")

```

#### Considering useful site metadata attributes

A metadata table can be very useful in the interpretation of field data. Creating a table adds to the rigor of data collection; this drives a need for completeness. Deciding on attributes for data collection comes from the research design.

Any metadata attribute, such as the elevation, should be concise.

#### Adding a row from the site metadata

```{r}

## Add the attribute line below the site numbers
gt_tbl <- bb_gt_with_site_attribute(
  bb_ordered,
  site_metadata = site_metadata,
  attr = "elevation",
  attr_label = "Elevation",
  title = "Ordered Braun–Blanquet table with elevation",
  subtitle = "Annotated presentation",
  blank_zeros = TRUE
)

gt_tbl



```

#### Add color

Overflow: By default, values outside the abundance scale are rendered in white. This prevents numeric site attributes (such as elevation) from being colored inadvertently. Users who wish to flag out-of-range values explicitly can change the overflow color for diagnostic purposes.

```{r}

## Add color to the ordered table
gt_tbl2 <- bb_gt_color(gt_tbl, max_val = 7)

## Show the table
gt_tbl2

```

### Presence/Absence Tables

In some studies, species data are recorded as presence/absence rather than as cover or abundance. This may reflect the nature of the data (e.g., rapid surveys, incomplete abundance estimates) or a deliberate analytical choice. Braun–Blanquet tables can accommodate this representation naturally.

In a presence/absence table, nonzero values indicate presence and zeros indicate absence. The analytical workflow—conversion to a Braun–Blanquet table format, ordering, structural metrics, and annotation—remains unchanged. What differs is how the table is presented.

#### Presenting presence/absence data

When values are binary, repeated display of the numeric value (typically 1) adds little information and can distract from structure. In such cases, it is often clearer to suppress the numeric symbol and allow table structure or color to carry the signal of presence.

This can be achieved at the presentation layer by replacing the value 1 with a symbol (such as X) or by suppressing it entirely:

#### Preparing and ordering the data

```{r}
library(readr)
library(gt)
library(bbstruct)

## Read the data table
options(readr.show_col_types = FALSE)

raw_data <- readr::read_csv(file=
        "species,  S01,  S02,  S03
         sp_A,       1,        0,        1
         sp_B,       0,        1,        0
         sp_C,       1,        1,        1")

## Print the table
gt(raw_data) |>
  tab_header(title="Presence/Absence", 
             subtitle="Raw Data Table")

## Convert to Braun-Blanquet table format
bb_table <- bb_from_raw_data(raw_data)

## Reorder the table (baseline view)
bb_ordered <- bb_reorder(bb_table, method = "blind")

## Get the table for the ordered BB data
gt_tbl <- bb_gt(bb_ordered, 
                title = "Ordered Presence/Absence Table", 
                blank_zeros = TRUE)

gt_tbl
```

#### Changing the cell symbols

Adding some code when using the gt table function changes the symbolism in the table cells. This new representation is typical of presence/absence tables.

```{r}

## Get the table for the ordered BB data
gt_tbl <- bb_gt(bb_ordered, 
                title = "Ordered Presence/Absence Table", 
                blank_zeros = FALSE)

## Replace the numbers with symbols
gt_tbl2 <- gt_tbl |>
  gt::fmt(
    columns = gt::everything(),
    fns = function(x) {
      x_chr <- as.character(x)
      x_chr[x_chr == "1"] <- "X"  ## or "✓"
      x_chr[x_chr == "0"] <- "•"
      x_chr
    }
  )

## Show the table
gt_tbl2

```

When color is applied, the presence of a species can then be indicated visually by the shaded cell alone, while absence remains blank.

#### Color and presence/absence tables

Color is particularly effective for presence/absence tables because it emphasizes patterns of co-occurrence without implying gradations of abundance. In this context, color should be used sparingly and consistently, with a fixed scale and neutral defaults so that presence is visible without overwhelming the table.

As with all enhancements described in this chapter, color does not alter the analysis. It provides an additional visual cue that supports inspection and interpretation.

```{r}

## Get the table for the ordered BB data
gt_tbl <- bb_gt(bb_ordered, 
                title = "Ordered Presence/Absence Table", 
                blank_zeros = TRUE)

## Replace the numbers with symbols
gt_tbl2 <- gt_tbl |>
  gt::fmt(
    columns = gt::everything(),
    fns = function(x) {
      x_chr <- as.character(x)
      x_chr[x_chr == "1"] <- " "  ## or "✓"
      x_chr[x_chr == "0"] <- ""
      x_chr
    }
  )

## Color the table
gt_tbl2 <- gt_tbl2 |> 
  bb_gt_color(max_val = 1, high="forestgreen")

## Print the table
gt_tbl2

```

### Relation to Braun–Blanquet abundance scales

Presence/absence tables can be viewed as a simplification of Braun–Blanquet abundance tables. In some workflows, presence/absence data may later be replaced or augmented by ordinal abundance classes (e.g., r, +, 1–5). The formatting approach demonstrated here—mapping symbols or suppressing numeric values at display time—extends naturally to those cases.

The choice between presence/absence and abundance representations should be guided by data quality and research goals. The tools presented here support both without requiring changes to the underlying workflow.

```{r}

options(readr.show_col_types = FALSE)

## Read the definitions
BB_scale <- readr::read_csv(file=
  'Value, Definition,    Range of Cover
   r,     Very rare,     "Solitary individual, insignificant cover"
   +,     Few / Sparse,  "Few individuals, small cover (< 5%)"
   1,     Plentiful,     "Many individuals, but still small cover (< 5%)"
   2,     Very Numerous,  5 – 25% cover (independent of abundance)
   3,     Any number,     25 – 50% cover
   4,     Any number,     50 – 75% cover
   5,     Any number,     75 – 100% cover')

## Print the table of definitions
gt(BB_scale)|>
  tab_header(title="Braun-Blanquet Scale")
  
```

#### Apply the Braun-Blanquet scale

The range of data in the following example run from 0 (absent) to 7 (75-100% cover). Once the BB table is ordered, the values in the table are reassigned to reflect the traditional Braun-Blanquet scale.

```{r}
options(readr.show_col_types = FALSE)

## Read the raw data 
raw_data <- readr::read_csv(file =
"species, S01, S02, S03, S04, S05, S06, S07, S08
 sp_A,    5,   7,   4,   2,   0,   0,   0,   0
 sp_B,    4,   6,   3,   0,   1,   0,   0,   0
 sp_C,    3,   0,   3,   4,   3,   0,   0,   0
 sp_D,    0,   3,   4,   6,   4,   3,   0,   0
 sp_E,    0,   0,   3,   5,   4,   2,   0,   0
 sp_F,    0,   2,   1,   2,   3,   0,   5,   0
 sp_G,    0,   0,   0,   0,   0,   3,   0,   0
 sp_H,    0,   0,   0,   0,   0,   3,   3,   4"
)

## Inspect raw data
gt(raw_data)

## Convert to Braun-Blanquet table format
bb_table <- bb_from_raw_data(raw_data)

## Reorder the table (baseline view)
bb_ordered <- bb_reorder(bb_table, method = "blind")

## Get the table for the ordered BB data
gt_tbl <- bb_gt(bb_ordered, 
                title = "Ordered table", 
                blank_zeros = TRUE)

## Rescale the value to the BB symbols
gt_tbl2 <- gt_tbl |>
  gt::fmt(
    columns = gt::everything(),
    fns = function(x) {
      x_chr <- as.character(x)
      x_chr[x_chr == "1"] <- "r"
      x_chr[x_chr == "2"] <- "+"
      x_chr[x_chr == "3"] <- "1"
      x_chr[x_chr == "4"] <- "2"
      x_chr[x_chr == "5"] <- "3"
      x_chr[x_chr == "6"] <- "4"
      x_chr[x_chr == "7"] <- "5"
      x_chr
    }
  )

gt_tbl2
```
