---
title: "Contours"
format: html
editor: visual
---

# Contours on Google Earth

## Introduction

The purpose is to interpret data points to a surface as a contour diagram. This diagram is then placed over the Google Earth landscape.

Google Earth provides some needed services for visualizations. Perhaps most important is the terrain model of the surface of the planet. Placing data on the terrain often gives important insight.

Contour diagrams are a way to visualize interpolated data.

## Start with a Contour Diagram

Here is a tiny example that works.

```{r}
# Install if needed: install.packages("akima")
library(ggplot2)
library(readr)
library(akima)

## Get the data.
## Data must have lat, lon, value columns.
data <- read_csv(col_names = TRUE, file =
 "lat,      lon,       value, id
  33.93006, -116.98778, 1627, P1
  34.33064, -117.32223, 1344, P2
  34.02718, -117.41113,  828, P3
  34.40641, -116.80974, 1625, P4
  34.45237, -117.38530,  876, P5
  33.73644, -117.29549,  529, P6
  34.12248, -117.57508, 1087, P7
  34.41393, -117.48995,  944, P8
  34.14114, -118.00842, 1575, P9
  34.06529, -118.24990,  627, P10
  34.46546, -116.86285, 1742, P11
  34.06266, -116.96609, 2707, P12
  34.24205, -117.32580, 1666, P13
  34.15810, -117.14770, 2860, P14
  33.78233, -118.45815,  647, P15
  34.41985, -117.68774, 1310, P16
  33.89687, -117.21061, 1398, P17
  33.73364, -118.13210,  458, P18
  33.96233, -117.95909,  753, P19
  34.46360, -118.10623, 1778, P20")  

## Interpolate onto a regular grid.
## Make 100x100 grid covering the lon & lat extents of the data.
grid_res <- 100
interp_result <- with(data,
  akima::interp(
    x = lon,
    y = lat,
    z = value,
    xo = seq(min(lon), max(lon), length.out = grid_res),
    yo = seq(min(lat), max(lat), length.out = grid_res),
    linear = TRUE,     # linear or spline interpolation
    extrap = FALSE     # Don't extrapolate beyond data range
  )
)

## Convert interpolation result to a data frame for ggplot.
interp_data <- data.frame(
  lon   = rep(interp_result$x, each = grid_res),
  lat   = rep(interp_result$y, times = grid_res),
  value = as.vector(interp_result$z)
)

## Remove any NAs (locations that couldn't be interpolated).
interp_data <- na.omit(interp_data)

## Plot the interpolated data as a filled contour diagram with a point overlay
combined_plot <- ggplot(interp_data, aes(x = lon, y = lat, z = value)) +
  geom_contour_filled(bins = 10) +
  geom_point(data = data, aes(x = lon, y = lat, z = value),size = 2, color = "red") +
  coord_fixed() +
  labs(title = "Interpolated Filled Contour",
       x = "Longitude", y = "Latitude")

print(combined_plot)

```

## Create Overlays for Google Earth

Here, we'll switch to the contour_to_earth function in the geContour package.

### Demo

```{r}
# Demo of the improved contour-to-earth package with Akima interpolation
# Source the improved contour-to-earth functions
## source("contour_to_earth.R")

## Load geContour.
install_github("kimbridges/geContour")
library(geContour)  ## Has the contour_to_google_earth function

# Load required libraries
library(ggplot2)

# Load or generate the sparse data
if (file.exists("socal_sparse_data.csv")) {
  socal_data <- read.csv("socal_sparse_data.csv")
  ## Confirm data
  gt(socal_data)|>
  tab_source_note(source_note="Test data")
} else {
  # Generate sample sparse data (20 points)
  set.seed(123)
  n_points <- 20
  
  # Generate random coordinates for Southern California
  socal_data <- data.frame(
    lat = runif(n_points, 33.7, 34.5),
    lon = runif(n_points, -118.5, -116.8)
  )
  
  # Generate values based on location (simulating elevation)
  socal_data$value <- sapply(1:nrow(socal_data), function(i) {
    lat <- socal_data$lat[i]
    lon <- socal_data$lon[i]
    
    # San Gabriel Mountains peak around 34.3, -118.0
    san_gabriel <- 2000 * exp(-((lat - 34.3)^2 + (lon - (-118.0))^2) * 15)
    
    # San Bernardino Mountains peak around 34.2, -117.0
    san_bernardino <- 2500 * exp(-((lat - 34.2)^2 + (lon - (-117.0))^2) * 12)
    
    # LA Basin depression around 34.0, -118.25
    la_basin <- -800 * exp(-((lat - 34.0)^2 + (lon - (-118.25))^2) * 10)
    
    # Add some random noise
    noise <- rnorm(1, mean = 0, sd = 200)
    
    # Combine all features
    elevation <- san_gabriel + san_bernardino + la_basin + noise
    
    # Normalize to a reasonable range
    return(max(0, round(elevation + 800)))
  })
  
  # Add point IDs
  socal_data$id <- paste0("P", 1:nrow(socal_data))
}

# Print the sparse data
cat("Using sparse data with", nrow(socal_data), "points:\n")
print(head(socal_data))

# Display the data distribution
cat("\nData distribution:\n")
cat("  Latitude range:", min(socal_data$lat), "to", max(socal_data$lat), "\n")
cat("  Longitude range:", min(socal_data$lon), "to", max(socal_data$lon), "\n")
cat("  Value range:", min(socal_data$value), "to", max(socal_data$value), "\n")

# Create a scatter plot to visualize the raw data points
scatter_plot <- ggplot(socal_data, aes(x = lon, y = lat, color = value)) +
  geom_point(size = 3) +
  geom_text(aes(label = id), hjust = -0.3, vjust = -0.3, size = 3) +
  scale_color_viridis_c() +
  coord_fixed() +
  theme_minimal() +
  labs(
    title = "Raw Data Points",
    subtitle = paste(nrow(socal_data), "scattered points in Southern California"),
    x = "Longitude",
    y = "Latitude",
    color = "Value"
  )

# Display the scatter plot
print(scatter_plot)

# Create a comparison of different interpolation methods

# 1. No interpolation (raw data)
raw_contour <- create_contour_plot(
  data = socal_data,
  contour_breaks = 8,
  interpolate = FALSE,
  title = "No Interpolation (Raw Data)"
)

# 2. Akima interpolation (default)
akima_contour <- create_contour_plot(
  data = socal_data,
  contour_breaks = 8,
  interpolate = TRUE,
  interp_method = "akima",
  interp_grid_res = 100,
  title = "Akima Interpolation"
)

# 3. Thin Plate Spline interpolation (TPS)
tps_contour <- create_contour_plot(
  data = socal_data,
  contour_breaks = 8,
  interpolate = TRUE,
  interp_method = "tps",
  interp_grid_res = 100,
  title = "Thin Plate Spline Interpolation"
)

# Display the interpolation comparison
print(raw_contour)
print(akima_contour)
print(tps_contour)

# Create Google Earth overlays with the improved functions

# Using Akima interpolation (recommended for sparse data)
akima_result <- contour_to_google_earth(
  data = socal_data,
  output_name = "socal_akima_contour",
  contour_breaks = 8,
  alpha = 0.7,
  buffer_percent = 1,
  include_legend = TRUE,
  create_kmz = TRUE,
  interpolate = TRUE,
  interp_method = "akima",
  interp_grid_res = 100
)

# Using Thin Plate Spline interpolation (alternative)
tps_result <- contour_to_google_earth(
  data = socal_data,
  output_name = "socal_tps_contour",
  contour_breaks = 8,
  alpha = 0.7,
  buffer_percent = 1,
  include_legend = TRUE,
  create_kmz = TRUE,
  interpolate = TRUE,
  interp_method = "tps",
  interp_grid_res = 100
)

# Create a custom color palette that might be suitable for elevation
terrain_colors <- colorRampPalette(c("darkgreen", "green", "greenyellow", 
                                    "yellow", "orange", "red", "darkred"))(8)

# Create a version with a custom color palette (good for elevation data)
custom_result <- contour_to_google_earth(
  data = socal_data,
  output_name = "socal_custom_colors",
  contour_breaks = 8,
  fill_colors = terrain_colors,
  alpha = 0.7,
  buffer_percent = 1,
  include_legend = TRUE,
  create_kmz = TRUE,
  interpolate = TRUE,
  interp_method = "akima"
)

# Display results and instructions
cat("\nGoogle Earth overlay files created:\n")
cat("\n1. Akima Interpolation:\n")
cat("   - KML:", akima_result$kml, "\n")
cat("   - KMZ:", akima_result$kmz, "\n")
cat("\n2. Thin Plate Spline Interpolation:\n")
cat("   - KML:", tps_result$kml, "\n")
cat("   - KMZ:", tps_result$kmz, "\n")
cat("\n3. Custom Color Palette (Akima Interpolation):\n")
cat("   - KML:", custom_result$kml, "\n")
cat("   - KMZ:", custom_result$kmz, "\n")

cat("\nTo use in Google Earth Pro:\n")
cat("1. Open Google Earth Pro\n")
cat("2. Go to File > Open...\n")
cat("3. Select any of the KMZ files to view the contour overlay\n")
cat("4. For best results with sparse data, use the Akima interpolation version\n")

```
