---
title: "Data"
format: html
editor: visual
---

## Image Data

The basic data consists of a set of photos that have identifying information. In large part, this is information that you add to the images. The data are stored in the JPG metadata in EXIF fields.

Adding information to the photos sounds like a lot of work. What we're looking at here are ways to make this relatively easy. There are several strategies so you have some options that best meet your situation.

There are basic data fields that are used. Don't worry about the names right now. The names correspond to EXIF fields and we're using those fields as standards the software understands.

The required data fields:

-   **Event:** This is used to *categorize* the images. It should be a short (usually one or two words) name. Multiple images will share this name. There are likely no more than 20 names for a set of images.

-   **Description:** This is a long text string that describes the specific image. Each image may have a different description.

-   **Location:** A specific place where the image was taken. Sometimes, all the images may be at the same place, or there can be a few places.

-   **City:** Like location, this is likely a shared value.

-   **State:** This is likely to be the same for all images.

-   **Country:** This is likely to be the same for all images.

-   **Latitude:** A specific geographic coordinate value, likely in decimal degrees. This may be shared with some or all the images.

-   **Longitude:** A specific geographic coordinate value, likely in decimal degrees. This may be shared with some or all the images.

## Example: A Foodie's Adventure

You've gone on vacation and had a great time visiting restaurants and eating a variety of meals. You want a photo book that reminds you of this experience. It's your habit to take a few photos of your meal at each of the restaurants.

### Starting with Photos

Let's assume that you've placed a set of food photos in a folder named "Photos". You've got 20 images of your meals. None of these images has the EXIF data (those fields listed above) with the required information. The fields are blank.

We're assuming that you leave the names of the photos (e.g., A002431.jpg) the same as they came out of the camera and are stored in a computer folder.

### Creating a List of Image Names

The file name is what we'll use as the image name. We can get a list by processing the contents of the file. We'll number this list in a way that it corresponds to the sorted order of the file names.

```{r}



```

### Adding Events to Categorize the Photos

Events, in our example, are the restaurants. Let's assume that you visited five restaurants. You're going to use between 3 and 6 images to show your meal at these places.

Let's use the numbered list of images and identify the range of images that corresponds to each restaurant. Our data are simply the start and end of the range and the name of the restaurant. A few restaurant names are abbreviated for convenience. Later, you'll see we will show the full names.

```{r}



```

### Describing the Food in Each Photo

Here, we're going to provide a few alternatives.

**Typing lines:** Each description is typed. The lines are in the order of the photos. This seems like the simplest way, but it may become tedious quite quickly.

**Dictating lines:** There are ways you can speak and have the computer type your words. In some cases, such as Google Docs, you can create a file by speaking and then copy-and-paste the text into your code. This can be quite efficient as you can look at a photo and speak as you see and remember what's in the image.

**AI image recognition:** This is, most likely, a two step procedure. You can use AI to do a first draft. Then you can edit the text to improve the quality of the description. This may work better than you expect for some situations, such as food. Other situation may require that you type or dictate the description as it's not obvious the intent of the image from what's in the photo. A potential limitation of this method is the need for an API key from the AI vendor (something that generally costs money).

#### Typing/Dictation

For our purposes, the appearance in the code is the same. Either you typed the lines or you dictated them and copied-and-pasted the text in the code.

Note that each line corresponds to a single image and that the order of the lines is the same as the order of the photos.

```{r}


```

#### AI Recognition

This is a three-step process. Once it is set up, it may be convenient to use.

1.  Image descriptions are generated by the AI engine and written to a file.

2.  The description lines in the file are edited. This isn't shown in the example below. This uses a simple text editor or even the editing capabilities in R Studio.

3.  The edited results are read back into this code where they can be used like the descriptions entered by typing or dictation.

#### Generate Description

```{r}


```

#### Read the Description

```{r}

```

### Adding the Location

Let's assume that the restaurants were in several different parts of a city. Here is where you add this information.

You can use the grouping strategy shown earlier (see Adding Events).

```{r}

```

### Filling in City, State and Country

For this example, we're assuming that all the restaurants are in the same city, state and country. You'll see that if this is true, that there is just one value for a field for all the images, then it's quite straightforward to enter the values.

```{r}

```

### Getting Geographic Coordinates

Some cameras automatically add the latitude and longitude to each image. Cell phone cameras do this. Other cameras, such as DSLR or mirrorless cameras might not do this. Here, we'll tackle a few strategies for the cases where the images don't come with geographic coordinates.

Here are some options:

Google Maps:

Reverse Geocode:

Interpolate from a Track Log:
