---
title: "Creating Maps Using R"
subtitle: "Simplifying the Process"
title-block-banner: true
author: "K. W. Bridges"
date: last-modified
date-format: long
format: 
  html:
    code-fold: true
    code-summary: "Show the code chunk"
editor: visual
---

# Inset Maps

It is often necessary to show two maps, each at a different scale. One map gives a general orientation (e.g. a country perspective) and the other map provides a detailed view of a study site. Sometimes these maps can be combined in to a single graphic. 



## General Setup & Initialization

Load the needed libraries. Then activate the Google Maps API key. This key provides access to Google Maps. Since this is a pay service, the key is kept secret.

The key registration needs to be done only once for an RStudio session. This is why the process is kept in a separate R chunk so we don't keep re-registering as we refine the maps.

```{r "library"}
#| warning: false 
#| error: false 
#| message: false

## Libraries
library(readr)         ## Read data
library(ggmap)         ## Enhance and show maps
library(cowplot)       ## Join maps
library(ggplot2)       ## General graphics support
library(dplyr)         ## Data manipulation
library(gt)            ## Make tables
library(ggrepel)       ## Automatic replacement of labels
library(googlesheets4) ## Read Google sheets

## Test if Google Key is registered
if (!has_google_key()){

  ## Initialize Google Map key; the is stored in a Project directory.
  My_Key <- read_file("P://Hot/Workflow/Workflow/keys/Google_Maps_API_Key.txt")

  ## Register the Google Maps API Key.
  register_google(key = My_Key, account_type = "standard")
  } ## end Google Key test

```

### Some useful code

There are a few bits of code, such as the assignment of data to a variable, that make it easier to code.

This is also where some useful functions reside. The functions are described elsewhere, but are located in this chunk for convenience.

```{r "functions"}
#| warning: false
#| error: false
#| message: false

## Define a theme that removes the axis labels and 
## puts a border around the map
simple_black_box <- theme_void() +
              theme(panel.border = 
                      element_rect(color = "black", 
                                   fill=NA, 
                                   size=2))

#######################################################
## Google Maps style defaults
## Use these to hide map layers
no_poi        <- "feature:poi|element:labels|visibility:off"
no_transit    <- "feature:transit|element:labels|visibility:off"
no_water      <- "feature:water|element:labels|visibility:off"
no_road       <- "feature:road|element:labels|visibility:off"
no_landscape  <- "feature:landscape|element:labels|visibility:off"
no_administrative <- "feature:administrative|element:labels|visibility:off"
hide_all <- "feature:all|element:labels|visibility:off"

#######################################################
## Functions ##########################################

##### bb_square #######################################

bb_square <- function(bb){

  ## Make a square boundary box

  x_length <- max(c(abs(bb[[1]]),abs(bb[[3]]))) - min(c(abs(bb[[1]]),abs(bb[[3]])))
  y_length <- max(c(abs(bb[[2]]),abs(bb[[4]]))) - min(c(abs(bb[[2]]),abs(bb[[4]])))

  if (x_length < y_length){
    half_diff <- (y_length - x_length) / 2
    bb[[1]] <- bb[[1]] - half_diff
    bb[[3]] <- bb[[3]] + half_diff
  } else {
    half_diff <- (x_length - y_length) / 2
    bb[[2]] <- bb[[2]] - half_diff
    bb[[4]] <- bb[[4]] + half_diff
  }

return(bb)
} ## end function bb_square

##### bb2google #######################################

bb2google <- function(bb){
  
  ## The bounding box can be produced using make_bbox.
  ## a bounding box has left, bottom, right, top coordinates.
  ## This function finds the map center and zoom level
  ## so these data can be used with a Google map request.
  
  ## This function assumes a square bounding box as input.
  
  ## Initialize
  info <- NULL

  ## Find the center point
  center_lon <- ((bb[[3]]-bb[[1]])/2) + bb[[1]]
  center_lat <- ((bb[[4]]-bb[[2]])/2) + bb[[2]]
  ##info$center <- c(lon=center_lon, lat=center_lat)
  info$lon <- center_lon
  info$lat <- center_lat

  ## Use the zoom scales to estimate the proper zoom level
  zoom_scales <- c(180,90,45,22.5,11.25,5.625,2.813,1.406,0.703,0.352,0.176,0.088,0.044,0.022,0.011,0.005,0.003,0.001,0.0005)

  bb_length <- bb[[3]] - bb[[1]]
  info$zoom <- as.integer(which.min(zoom_scales > bb_length))

  return(info) 

}  ## end function bb2google

```

## Place an Overlay on a Country

An ovelay makes the location and extent of a country stand out from the neighboring areas. This can be especially useful for general orientation.

The `ggplot2` package includes some mapping functions. One of these is `map_data.` Use this to return an outline of a country. Other administrative levels, such as US States, can be used. That will be shown later.

The map_data function not only returns the country outline, it also provides geographic coordinates that enable the overlay registration. These coordinates are called `long` and `lat` (a slight departure from the usual convention of `lon` and `lat`).

```{r "combine"}
#| warning: false
#| error: false
#| message: false
#| results: false

## Specify the area
country <- "Costa Rica"
zoom    <- 6

## Get the Google Map which will be the basemap.
basemap <- get_googlemap(center=country,
                         zoom=zoom,
                         maptype = "terrain",
                         scale=2,
                         style = hide_all) 

## Get the overlay map that will highlight the area of interest.
overlay <- map_data(map="world", region=country, interior = TRUE) 

## Combine the two maps. Note: This source uses long,lat from
## the map_data function and are part of the overlay data frame.
combo_map <- ggmap(basemap) +
  geom_polygon(data = overlay,
               aes(x=long,y=lat,group=group),
                 color = "black", ## Outline color
                 line  = 2,       ## Outline thickness
                 fill  = "wheat", ## Overlay color
                 alpha = 0.5) +   ## Make it a "see through" overlay
              labs(x="",y="") +
              simple_black_box

## Display the map.
combo_map

```


## Building a Basemap with an Inset Locator Map


```{r}
#| warning: false
#| error: false
#| message: false

## Get the data from a Google Sheet.
bot_gardens <- range_speedread("1ZhRkINFnbqWknQJAXx7tA_yW1NgPUW_Ssr7MR2rI1CM",
                               sheet = "prelim",
                               show_col_types = FALSE)

## Print a table for a quick data check
gt(bot_gardens) %>% 
  fmt_number(columns = c(2,3),decimals = 5)

## Make sure the column names are all lowercase and separate out the data points
names(bot_gardens) <- tolower(names(bot_gardens))
garden_points <- bot_gardens %>% 
  dplyr::select(lon,lat)

## Build a square bounding box from the data points and get info for a Google Map
garden_bb <- make_bbox(lon=garden_points$lon, lat=garden_points$lat, f=0.1)
garden_square <- bb_square(garden_bb)
garden_google <- bb2google(garden_square)

## Create a Stamen basemap with data points
## First get the Stamen map
garden_map <- get_stamenmap(bbox = garden_square,
                            maptype="terrain")
                                 
basemap <- ggmap(garden_map) + 
    simple_black_box +
    geom_point(aes(x = lon, y = lat, color = "red", size=1.0), 
               data = garden_points) +
    theme(legend.position = "none", aspect.ratio = 1.0) ## help prevent distortion

## Get the Stamen map which will be the Inset map.
insetmap <- get_map(location="Japan",
                    zoom=6,
                    source="stamen",
                    maptype="terrain-background")

## Create the Inset Map with a red box showing the area of the basemap
inset <- ggmap(insetmap) + 
    simple_black_box +
    geom_rect(aes(xmin=garden_square[[1]],ymin=garden_square[[2]],xmax=garden_square[[3]],ymax=garden_square[[4]]),
              color = "red",
              size  = 1.0,
              fill  = NA)

## Combine the two maps
gg_inset <- ggdraw() +
  draw_plot(basemap) +
  draw_plot(inset, x= 0.14, y=0.63, 
            width = 0.35, height = 0.35, 
            hjust=0, vjust = 0)
 
## Show the combined maps
gg_inset

```

