---
title: "Creating Maps Using R"
subtitle: "Simplifying the Process"
title-block-banner: true
author: "K. W. Bridges"
date: last-modified
date-format: long
format:
  html:
    fig-width: 8
    fig-height: 8
    code-fold: true
    code-summary: "Show the code chunk"
editor: visual
execute:
  warning: false
  error: false
  message: false
---

# Basemap Gallery

Getting an appropriate basemap is an essential first step in the map making process. The examples shown in this gallery show some of the variety you can get by using fairly simple R code.

Later, you'll see that the specification of a basemap is made easier by using some functions. Here, we're only focused on the general appearance of the basemaps. These examples will help you decide what type of basemap will best suit your project.

## Setup & Defaults

::: callout-important
## Setup Information

Each section has its own setup and default information. This allows each section to be run independently. The duplication of code is a small price to be paid for the independence and clarity achieved by this repetition.
:::

All use of Google maps requires the use of an API key. Getting this key is a one-time task. Store the key somewhere same and hidden from the outside.

Registering the Google Key needs to be done once each session.

```{r "library"}

## Libraries
library(readr)
library(ggmap)
library(ggplot2)
library(sitemaps)

## Initialize Google Map key; the is stored in a Project directory.
  My_Key <- read_file("P://Hot/Workflow/Workflow/keys/Google_Maps_API_Key.txt")

## Test if Google Key is registered
if (!has_google_key()){

  ## Register the Google Maps API Key.
  register_google(key = My_Key, account_type = "standard")
  } ## end Google Key test

```

### Some Required and Useful Code

It is a good practice to include the code to set the parameters (`set_param()`) and the Google Map hide codes (`set_google_hides()`).

A simple `theme` is defined here that removes the geographic coordinates from the map borders, places a black line around the map, and suppresses any legend.

```{r "required"}

## Read in all the parameters
column <- site_styles()
hide   <- site_google_hides()

#################################################################

## Define a theme that removes the axis labels and 
## puts a border around the map. No legend.
simple_black_box <- theme_void() +
              theme(panel.border = element_rect(color = "black", 
                                   fill=NA, 
                                   size=2),
                    legend.position = "none")

```

## Google Basemaps

The **Google** basemaps have annotations on top. There are ways to selectively limit the annotation. This capability is discussed later along with a few other features that make these maps more useful.

|              |                                |
|--------------|--------------------------------|
| **Map Type** | **Appearance**                 |
| roadmap      | Street map                     |
| terrain      | Shaded relief map              |
| satellite    | Satellite image                |
| hybrid       | Satellite image and street map |

Here are a few examples that show some of the types of base maps. Note that the theme ("`simple_black_box`"), defined earlier, is used to simplify these maps.

```{r "google1"}
#| label: fig-google1
#| fig-cap: "Oahu, Hawaii as a roadmap."

map2 <- get_map(location = "Island of Oahu, Hawaii, USA", 
                zoom=10, 
                source="google",
                maptype="roadmap")

ggmap(map2) + simple_black_box

```

```{r "google2"}
#| label: fig-google2
#| fig-cap: "Oahu, Hawaii as a terrain map."

map3 <- get_map(location = "Island of Oahu, Hawaii, USA", 
                zoom=10, 
                source="google",
                maptype="terrain")

ggmap(map3) + simple_black_box

```

```{r "google3"}
#| label: fig-google3
#| fig-cap: "Oahu, Hawaii as a satellite map."

map5 <- get_map(location = "Island of Oahu, Hawaii, USA", 
                zoom=10, 
                source="google",
                maptype="satellite")

ggmap(map5) + simple_black_box

```

```{r "google4"}
#| label: fig-google4
#| fig-cap: "Oahu, Hawaii as a hybrid map."

map1 <- get_map(location = "Island of Oahu, Hawaii, USA", 
                zoom=10, 
                source="google",
                maptype="hybrid")

ggmap(map1) + simple_black_box

```

### Google Simplifications

It is possible to "hide" different layers of annotations on Google Maps. This example, which shows the `hide$all` parameter is just one possibility. See Chapter 3 (Google Maps Simplified) for examples of alternative layer hiding.

```{r "gsimple"}
#| label: fig-gsimple
#| fig-cap: "Oahu, Hawaii as a terrain map with layers hidden."
#| 
map <- get_googlemap(center = "Island of Oahu, Hawaii, USA", 
                zoom=10, 
                maptype="terrain",
                style=hide$all)

ggmap(map) + simple_black_box

```

#### Scale Differences

The large property shown in this example houses the Huntington Library, Art Gallery and Botanical Garden. Zooming out, you see the surrounding neighborhood and, with even more zoom, the city is shown next to the adjacent mountains.

```{r "scale1"}
#| label: fig-scale1
#| fig-cap: "Southern California at zoom = 16 on a terrain map."

map10 <- get_map(location = c(lon=-118.11418, lat=34.12839),
                zoom=16,
                source="google",
                maptype="terrain")

ggmap(map10) + simple_black_box

```

```{r "scale2"}
#| label: fig-scale2
#| fig-cap: "Southern California at zoom = 14 on a terrain map."

map11 <- get_map(location = c(lon=-118.11418, lat=34.12839),
                zoom=14,
                source="google",
                maptype="terrain")

ggmap(map11) + simple_black_box

```

```{r "scale3"}
#| label: fig-scale3
#| fig-cap: "Southern California at zoom = 12 on a terrain map."

map12 <- get_map(location = c(lon=-118.11418, lat=34.12839),
                zoom=12,
                source="google",
                maptype="terrain")

ggmap(map12) + simple_black_box

```

```{r "scale4"}
#| label: fig-scale4
#| fig-cap: "Southern California at zoom = 10 on a terrain map."

map13 <- get_map(location = c(lon=-118.11418, lat=34.12839),
                zoom=10,
                source="google",
                maptype="terrain")

ggmap(map13) + simple_black_box

```

## Stamen Basemaps

Similarly, the **Stamen maps** have very different appearances. Some of the basemap alternatives, such as "watercolor," are very creative and might be interesting to use for non-technical applications.

In practice, Stamen maps of large areas may involve considerable downloading. It is suggested that they only be used for smaller areas.

| Map Type           | Appearance                                                   |
|------------------|-----------------------------------------------------|
| terrain-background | Terrain without names                                        |
| terrain            | Mountains with hill-shading, colored vegetation, and names   |
| toner              | Black & white, minimalist without names                      |
| toner-background   | Black & white, minimalist without names                      |
| toner-lite         | Grey & white, minimalist with names                          |
| watercolor         | Colorful, abstract, without streets, place names, or borders |

```{r "simple1"}
#| label: fig-simple1
#| fig-cap: "Japan with a Stamen terrain-background map."

map8 <- get_map(location="Japan",
                zoom=5,
                source="stamen",
                maptype="terrain-background")

ggmap(map8) + simple_black_box

```

```{r "simple2"}
#| label: fig-simple2
#| fig-cap: "Japan with a Stamen terrain map."

map9 <- get_map(location="Japan",
                zoom=5,
                source="stamen",
                maptype="terrain")

ggmap(map9) + simple_black_box

```

```{r "simple3"}
#| label: fig-simple3
#| fig-cap: "Japan with a Stamen toner map."

map12 <- get_map(location="Japan",
                zoom=5,
                source="stamen",
                maptype="toner")

ggmap(map12) + simple_black_box

```

```{r "simple4"}
#| label: fig-simple4
#| fig-cap: "Japan with a Stamen toner-background map."

map13 <- get_map(location="Japan",
                zoom=5,
                source="stamen",
                maptype="toner-background")

ggmap(map13) + simple_black_box

```

```{r "simple5"}
#| label: fig-simple5
#| fig-cap: "Japan with a Stamen terrain-lite map."

map10 <- get_map(location="Japan",
                zoom=5,
                source="stamen",
                maptype="toner-lite")

ggmap(map10) + simple_black_box

```

```{r "simple6"}
#| label: fig-simple6
#| fig-cap: "Japan with a Stamen watercolor map."
map14 <- get_map(location="Japan",
                zoom=5,
                source="stamen",
                maptype="watercolor")

ggmap(map14) + simple_black_box

```

These maps give a lot of options when you're choosing a basemap.

## Adding Place Names (preview)

Maps that don't have names can be given the place names you want. This is an important part of the process in creating a custom map that best serves your visualization requirements.

There is a detailed treatment giving names in a later chapter (Chapter 6: Names). Here is a simple demonstration that uses the last map (

You can see (in the code) that the names are provided in a table along with coordinate locations and a few style parameters. How these column values work will be explained later.

```{r "namespreview"}
#| label: fig-namespreview
#| fig-cap: "Japan with a Stamen watercolor map with place names."

map14names <- read_csv(col_names = TRUE, file = 
    "text,          lat,      lon,       name_text_wrap, fontface
     Tokyo,         35.68668, 139.77796, NA,             plain
     Kyoto,         35.01866, 135.76851, NA,             plain
     Hiroshima,     34.45604, 132.47520, NA,             plain
     Sapporo,       43.14579, 141.34479, NA,             plain
     Sea of Japan,  39.94473, 134.04759, 9,              italic
     Pacific Ocean, 30.91643, 140.34591, 9,              italic")

ggmap(map14) +
  site_names(datatable = map14names) +
  simple_black_box

```

```{r "calif1"}

map10 <- get_map(location="California, USA",
                zoom=6,
                source="stamen",
                maptype="toner")

ggmap(map10) + simple_black_box

map11 <- get_map(location="California, USA",
                zoom=6,
                source="stamen",
                maptype="toner-lite")

ggmap(map11) + simple_black_box

map12 <- get_map(location="California, USA",
                zoom=6,
                source="stamen",
                maptype="toner-background")

ggmap(map12) + simple_black_box

```
