---
title: "Univariate"
format: html
---

# Univariate Analysis

### Data Distribution

Prompt

I need R code to analyze a set of numbers. Specifically, I need to determine the basic statistics (e.g., mean, median) and determine which is better based on whether the data are normally distributed or not. With this decision, I then need to see the statistical values appropriate for the better descriptor and a chart that shows the data in the context of the proper distribution.

***The boxplot doesn't look right for univariate data.***

```{r}
# Load required libraries
library(ggplot2)

# Function to analyze data
analyze_data <- function(data) {
  # Basic statistics
  mean_val <- mean(data)
  median_val <- median(data)
  sd_val <- sd(data)
  
  # Test for normality
  shapiro_test <- shapiro.test(data)
  is_normal <- shapiro_test$p.value > 0.05
  
  # Determine better descriptor
  better_descriptor <- ifelse(is_normal, "Mean", "Median")
  
  # Calculate additional statistics
  if (is_normal) {
    additional_stats <- c(mean = mean_val, sd = sd_val)
  } else {
    additional_stats <- c(median = median_val, 
                          q1 = quantile(data, 0.25), 
                          q3 = quantile(data, 0.75))
  }
  
  # Create plot
  if (is_normal) {
    p <- ggplot(data.frame(x = data), aes(x)) +
      geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
      geom_density(color = "red") +
      geom_vline(xintercept = mean_val, color = "blue", linetype = "dashed") +
      labs(title = "Data Distribution with Normal Curve",
           x = "Value", y = "Density") +
      theme_minimal()
  } else {
    p <- ggplot(data.frame(x = data), aes(x)) +
      geom_boxplot(fill = "lightblue", color = "black") +
      labs(title = "Data Distribution (Box Plot)",
           x = "Value") +
      theme_minimal()
  }
  
  # Return results
  list(
    basic_stats = c(mean = mean_val, median = median_val),
    is_normal = is_normal,
    better_descriptor = better_descriptor,
    additional_stats = additional_stats,
    plot = p
  )
}

# Example usage
set.seed(123)  # for reproducibility
data <- rnorm(100, mean = 10, sd = 2)  # replace with your actual data

results <- analyze_data(data)

# Print results
print("Basic Statistics:")
print(results$basic_stats)
print(paste("Data is normally distributed:", results$is_normal))
print(paste("Better descriptor:", results$better_descriptor))
print("Additional Statistics:")
print(results$additional_stats)

# Display plot
print(results$plot)
```

Sample Size

Prompt

Can I see some code that will determine if I have sampled enough, given that I want to have at least 5% confidence that the mean is within 5% of the correct value?

```{r}
# Function to check if sample size is adequate
check_sample_size <- function(data, confidence_level = 0.95, margin_of_error_percent = 0.05) {
  # Calculate sample size and mean
  n <- length(data)
  sample_mean <- mean(data)
  
  # Calculate standard error
  se <- sd(data) / sqrt(n)
  
  # Calculate t-score for the given confidence level
  degrees_of_freedom <- n - 1
  t_score <- qt((1 + confidence_level) / 2, degrees_of_freedom)
  
  # Calculate margin of error
  margin_of_error <- t_score * se
  
  # Calculate the current margin of error as a percentage of the mean
  current_margin_percent <- margin_of_error / abs(sample_mean)
  
  # Check if the current margin of error is within the desired percentage
  is_adequate <- current_margin_percent <= margin_of_error_percent
  
  # Calculate required sample size if not adequate
  required_n <- ifelse(!is_adequate,
                       ceiling((t_score * sd(data) / (margin_of_error_percent * abs(sample_mean)))^2),
                       n)
  
  # Return results
  list(
    sample_size = n,
    is_adequate = is_adequate,
    current_margin_percent = current_margin_percent,
    required_sample_size = required_n
  )
}

# Example usage
set.seed(123)  # for reproducibility
data <- rnorm(15, mean = 10, sd = 2)  # replace with your actual data

result <- check_sample_size(data)

# Print results
cat("Current sample size:", result$sample_size, "\n")
cat("Is sample size adequate?", result$is_adequate, "\n")
cat("Current margin of error (as % of mean):", result$current_margin_percent * 100, "%\n")
if (!result$is_adequate) {
  cat("Required sample size for adequacy:", result$required_sample_size, "\n")
}
```
