# Packages

Several packages are used in the examples. These need to be installed to your local computer. Then each package needs to be activated using the `library()` function.

Most of the packages you will use are located on CRAN. Installing these packages can be done using RStudio's Install menu. The packages you will need are the following.

-   tidyverse

-   gt

-   devtools

## Installing the milestones package

The `milestones` package introduced in this document is not available on CRAN. Instead, it is easily downloaded from GitHub.

Before you can download `milestones` from GitHub, you need to activate the devtools package with a `library()` function. Then you can use the `install_github()` function to download the `milestones` package.

The following code chunk shows how to do this.

```{r}
#| label: install
#| warning: false
#| error: false

library(devtools)

install_github("kimbridges/milestones")

```

Once all the required packages are installed on the local computer, this task doesn't need to be done again.

Running the functions in the packages requires that each package is activated using the `library()` function. The following chunk activates all of the packages used in the examples.

```{r}
#| label: core

## core packages
library(tidyverse)
library(gt)

## specialty packages
library(devtools)
library(milestones)

```

## Using packages with chapters

It is necessary to put the activation code (i.e., library functions) near the start of each chapter.
