---
title: "QR Procedures"
author: "K. W. Bridges"
date: "9/2/2020"
output: pdf_document
---

```{r setup, include=FALSE,echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--
KEY HANDLING (added 2026-07-29)

This file used to carry a Google Maps / Elevation API key as a literal string in
two chunks. Both now read Sys.getenv("GGMAP_GOOGLE_API_KEY"), which is what
site_extract() and the vouchr package have always used.

Two things to know:

1. This document sets echo = TRUE globally and renders to PDF. Anything written
   as a literal in a chunk is printed into the rendered output. That is why a key
   must never be a literal here, even in a private folder.

2. Editing this file does NOT remove the old key from Google Drive revision
   history. Only revoking the key at the provider actually neutralises it.

The current derive stack is deliberately keyless -- site_extract() defaults to
source = "aws", and Macrostrat, SSURGO and Nominatim need no key either. The
Google path is a legacy fallback, so there may be nothing left that needs a key
at all.
-->

# QR-Codes: Generation and Scanning

This project has a set of chunks. Each chunk performs a specific task. Use the chunks one at a time to accomplish specific things.

DON'T RUN THE WHOLE SET AT ONE TIME! Choose what you need to do.

Chunk Directory

-   **Init-Stuff:** Run this to initialize all the libraries needed for the chunks.
-   **QR-Collector:** Creates images with collection IDs that are used as specimen labels.
-   **QR-Species:** Creates images with species names. Use these to with \*\*QR-ID-exercise to complete an identification project.
-   **Image--File-Listing:** Creates a list of file names for all images in a folder. Choose whether `.png` or `.jpg` files.
-   **Photo-EXIF:** Builds a list of the photo files in a folder.
-   **QR-Scan:** Initiates a scan, using Zbar, to find and decode the QR-Codes in a set of photos.
-   **XML-Decode:** Extracts the file and QR-Code from the XML file created by Zbar.

```{r Init-Stuff,echo=FALSE,warning=FALSE,error=FALSE}

## Always run this first!

library(qrcode)
library(grid)
library(gridExtra)
library(gridtext)
library(png)
library(exiftoolr)
library(gt)

## You need to run install_exiftool() from the Console

```

## Generate QR-Codes

The QR-Codes are created as images, with each images stored as a `.png` file in the base director. Images have a single line of text that is the same as the information in the QR-Code.

There are two types of image files produced, each one processed in a separate chunk. The intent, in both cases, is to provide a QR-Code that will be photographed so that the camera's GPS location can be linked to either a voucher specimen or a set of photographs of a species.

One set is for collector labels and its QR-Code contains the collector's ID and collection number. These images will eventually be printed on business-card stock, punched with a hole, and supplied with a short piece of string.

The other set if for photo-location species identification. The QR-Code has a species name (i.e., scientific name). In practice, each species on a site checklist will have a QR-Code printed on a sheet of paper. Other information will also be on the paper, such as the plant Family, Stature/Life Form, and Common Names. These QR-Codes are photographed to associate species with locations.

### Q-Collector: Create QR-Codes for Specimen Labels

The following information is supplied.

-   Collector name or code
-   Starting collection number
-   Number of labels needed
-   Name of a PDF file for the labels (each labels is a page in this file)

The chunk produces just a single PDF file. You print this file with a laser printer, specifying ten labels in two columns. Business card stock (such as Avery 5371)

```{r QR-Collector,echo=FALSE,warning=FALSE,error=FALSE}

## This version is for collector label QR-Codes.
## The chunk generates labeled QR codes as png files. These images are put in the base directory.

collector   <- "KWB:"   ## QR Code header
coll_number <- 1239     ## Starting number of the QR codes
QR_number   <- 10       ## The number of QR codes needed

last_coll_number <- coll_number + QR_number -1
PDF_file <- "Collection_labels.pdf" ## The PDF file for the labels

for (i in seq(coll_number, last_coll_number, 1)){

  code    <- paste(collector,i)
  QR_name <- paste0("temp_qr_",i,".png")

  png(QR_name)
  qrcode_gen(code)
  dev.off()

  img       <- rasterGrob(as.raster(readPNG(QR_name)), interpolate=FALSE)
  ID_number <- textGrob(code,gp=gpar(fontsize=48,font=2))

  png(QR_name)
  grid.arrange(img,top=ID_number)
  dev.off()
 
  }  ## End of loop for number of labels

## Put all the images in a single file as separate PDF pages

pdf(PDF_file,paper="USr")
for (i in coll_number:last_coll_number) {
    QR_name <- paste0("temp_qr_",i,".png")
    img <-  rasterGrob(as.raster(readPNG(QR_name)), interpolate=FALSE)
    grid.arrange(img)
    }
 dev.off()

## Get rid of the temporary files

sapply(paste0("temp_qr_", coll_number:last_coll_number, ".png"), unlink)
```

### QR-Species: Build Pages for Species Photography

This chunk creates content for a species-photography exercise.

The objective is to produce sheets of paper, each with a combination of a QR-Code with a species name and a set of ancillary information that helps in identifying the species in the field.

The input consists of these elements.

-   **Species Checklist:** Names of all the plant species found in the study area.
-   **Plant Information:**
-   **Plant Family**
-   **Species Code** (usually a four letter abbreviation of the genus and species)
-   **Common Names** (local names)
-   **Stature/Life Form**
-   **Distribution Status** (endemic, indigeneous, naturalized, invasive, etc.)

All these elements come from a `.csv` file, such as one created in a spreadsheet.

> Species, Family, SpeciesCode, CommonNames, Stature, Distribution

The output is a `.pdf` file for each species.

Combine all the `.pdf` files into a single file. Print this larger file. Pages are designed to be folded and laminated.

```{r QR-species,echo=FALSE,warning=FALSE,error=FALSE}

## This chunk creates species QR-Code images. The code is the species name.
## Input is a species list.

checklist <- read.csv(file="Checklist.csv", header=TRUE, stringsAsFactor = FALSE, fileEncoding="UTF-8")

species_number <- length(checklist)

## Create QR-Code for each species on the checklist
## Put the species name above the QR-Code

for (i in seq(1,species_number, 1)){

Spp_ID <- checklist$species[i]
QR_name <- paste0("Spp_QR_",i,".png")

png(QR_name)
qrcode_gen(Spp_ID)
dev.off()

img <-  rasterGrob(as.raster(readPNG(QR_name)), interpolate=FALSE)
Spp_info <- textGrob(Spp_ID,gp=gpar(fontsize=36,font=2))

png(QR_name)
grid.arrange(img,top=Spp_info)
dev.off()

}

## Create info listing from data about each species on the checklist

for (i in seq(1,species_number,1)){

line[1] <- paste("<i>",checklist$species[i],"</i><br>")
line[2] <- paste("Family:",checklist$family[i],"<br>")
line[3] <- paste("Abbreviation:",checklist$abbrev[i],"<br>")
line[4] <- paste("Common Name:",checklist$common[i],"<br>")
line[5] <- paste("Stature:",checklist$stature[i],"<br>")
line[6] <- paste("Distribution:",checklist$distribution[i])

Spp_string <- paste(line[1],line[2],line[3],line[4],line[5],line[6])

## This is a neat box with the information

g <- textbox_grob(
  Spp_string,
  x = unit(0.5, "npc"), y = unit(0.7, "npc"), halign = 0, valign = 1,
  gp = gpar(fontsize = 24),
  box_gp = gpar(col = "black", fill = "lightgray"),
  r = unit(5, "pt"),
  padding = unit(c(10, 10, 10, 10), "pt"),
  margin = unit(c(0, 10, 0, 10), "pt")
)
 # grid.newpage()  ## Debug: Prints the box
 # grid.draw(g)    ## Debug: Prints the box (check color, size)

## Output the QR-Code (top) and the species information (bottom) in separate PDF files (one for each species)

out_info <- paste0("Spp_Sheet_",i,".pdf")
QR_name <- paste0("Spp_QR_",i,".png")
img <-  rasterGrob(as.raster(readPNG(QR_name)), interpolate=FALSE)

pdf(out_info)
grid.arrange(img,g)
dev.off()

}
```

\newpage

## Image File Utilities

## Image-File-Listing: Create a List of Photo Files

Hint: It might be possible to run the Zbarimg commands using the following general structure:

shell.exec("\\\\network\\path\\file.bat")

The info can be sent to a file where it is picked up back in R.

\newpage

```{r Image-File-Listing,echo=FALSE,warning=FALSE,error=FALSE }

## This chunk generates a list of file names that end in "jpg" or "png" from all the files in the folder on the file path.
## The "files_list" can be used in other chunks.

files_path <- "Photos/Healer_species"
files_pattern <- "*.jpg"

files_list <- list.files(files_path,files_pattern)

length(files_list)
files_list

```

## Photo-EXIF: Obtain Location and Image Date

## Function to get the Google elevations

###### THIS IS NOT USED

```{r}

## <script src="https://gist.github.com/italo-goncalves/1a69523306d580ace8026fab17aec784.js"></script>
get_Google_Elevation <- function(lat, long, key){
        # packages
        require(httr)
        require(jsonlite)
        require(dplyr)
        
        # setup
        Ndata <- length(lat)
        base_int <- 40 # max roughly 40 data points per request
        Ncells <- ceiling(Ndata/base_int) 
        baseURL <- "https://maps.googleapis.com/maps/api/elevation/json?locations="
        
        # conversion to character vector
        lat_long_str <- character(Ncells)
        for (i in 1:(Ncells-1)){
                lat_long_str[i] <- 
                        paste(lat[((i-1)*base_int+1):(i*base_int)],
                              long[((i-1)*base_int+1):(i*base_int)],
                              sep = ",", collapse = "|")
        }
        lat_long_str[Ncells] <- paste(lat[-(1:(i*base_int))],
                                  long[-(1:(i*base_int))],
                                  sep = ",", collapse = "|")
        
        # request to Google Elevation API
        elev <- data.frame()
        for (i in 1:Ncells){
                # Sys.sleep(0.1)  # pause
                url <- paste0(baseURL,lat_long_str[i],"&key=",key)
                tryCatch({
                        json <- fromJSON(
                                content(GET(url), as = "text"), 
                                simplifyDataFrame = T)
                        cat("Request",i,"of",Ncells,
                            "- Status =",json$status,"\n")
                        if(json$status == "OK"){
                                gdata <- as.data.frame(json$results)
                                gdata <- tbl_df(data.frame(gdata$location, 
                                                           gdata$elevation, 
                                                           gdata$resolution))
                                elev <- bind_rows(elev, gdata)
                        }
                }, 
                error = function(e) cat("Error during request", i)
                )
        }
        
        # end
        colnames(elev) <- c("latitude","longitude","elevation","resolution")
        return(elev)
}
```

## Process as much as possible from photo info

```{r Photo-EXIF, echo=FALSE,warning=FALSE,error=FALSE}

library(lubridate)
library(ggmap)
library(googleway)
library(readtext)
library(httr)
library(dplyr)
library(jsonlite)
library(exiftoolr)

## This chunk reads EXIF data from JPG files. 
## Specifically, the location and date of the photo.

## This requires photos loaded from the camera
## (not saved from photos.google.com)

qr_photo   <- "IMG_20200126_092948_K.jpg"
photo_exif <- exif_read(qr_photo)

photo_exif$GPSLatitude
photo_exif$GPSLatitudeRef
photo_exif$GPSLongitude
photo_exif$GPSLongitudeRef
photo_exif$DateTimeOriginal

## Store the location in easy-to-use variables
in_lat <- photo_exif$GPSLatitude
in_lon <- photo_exif$GPSLongitude

in_lat_list <- list(in_lat)
in_lon_list <- list(in_lon)

in_location <- c(in_lon, in_lat)
in_loc_df <- data.frame(lat=in_lat,lon=in_lon)

## Process the date and time
in_date <- as.Date(substring(photo_exif$DateTimeOriginal,1,10),"%Y:%m:%d")

lub_str   <- ymd_hms(photo_exif$DateTimeOriginal) 
in_hour   <- hour(lub_str)
in_minute <- minute(lub_str)

## Get the name of the place
key4 <- Sys.getenv("GGMAP_GOOGLE_API_KEY")   ## 2026-07-29: was a hard-coded literal here.

##my_g_key <- as.data.frame(readtext("key/Google_Maps_API_Key.txt"))
register_google(key4)

place_name       <- revgeocode(in_location,output = "all")
loc_temp         <- place_name$results
in_location_name <- unname(unlist(loc_temp[1])[29])

## Site Elevation

##### Doesn't Like KEY


# set_key(key=key4)
# 
# google_keys()
# 
# google_map(location=in_location,zoom=6,key=my_g_key,width=4,height=4)
# 
# in_elevation <- google_elevation(in_loc_df,simplify = TRUE)
# in_elevation

#in_elevation <- get_Google_Elevation(lat = in_lat_list,long = in_lon_list, key = key4)

## NEW TRY -- from Google, using HTTR

library(httr)      ## JSON functions
library(jsonlite)  ## Decode JSON strings

in_lat <- format(round(in_lat,6), nsmall = 6)
in_lon <- format(round(in_lon,6), nsmall = 6)

##############################################################################

my_elevation_function <- function(lat,lon){

g_base <- "https://maps.googleapis.com/maps/api/elevation/json?"
g_key  <- Sys.getenv("GGMAP_GOOGLE_API_KEY")  ## 2026-07-29: was a hard-coded literal here.
g_locs <- paste0("locations=",lat,",",lon)

g_string <- paste0(g_base,g_locs,"&key=",g_key)

g_return <- httr::GET(g_string)
g_data   <- httr::content(g_return,"text")

## Resolution is the max distance from which the point elevation was interpolated

elev_data    <- jsonlite::fromJSON(g_data)  
elev_results <- elev_data$results
# elev_results$elevation
# elev_results$location
# elev_results$resolution

return(elev_results)
} ## end my_elevation_function

###################################################################################

```

## Slope

Make a 3 x 3 grid based on the lat,long point at the center.

Evaluate the resolution of each point to see if there is overlap of the max interpolation distances.

```{r}

library(geosphere) ## Spherical trigonometry
library(cimir)     ## To get compass points
library(raster)    ## Some of the raster calcs, like slope

## Flower Fields, generally west slope
place  <- "Flower Fields"
in_lon <- -117.319254
in_lat <- 33.128982
  
## Lombard Street, San Francisco, generally east facing slope
place  <- "Lombard Street"
in_lon <- -122.418957
in_lat <- 37.802111
  
## Mt Fuji, South facing slope
place  <- "Mt Fuji"
in_lon <- 138.730496
in_lat <- 35.355235

## Koko Crater Test Location
place  <- "Koko Crater"
in_lon <- -157.68687
in_lat <-   21.28164

```

```{r}

## Set a size (in decimal degrees) for the distance between points.
spacing <- 0.0002

## Add the spacing distance to the original latitude
in_lat_n <- as.numeric(in_lat)
in_lon_n <- as.numeric(in_lon)

## S-N (Bottom to Top)
lat1 <- in_lat_n - spacing
lat2 <- in_lat_n
lat3 <- in_lat_n + spacing

## W-E (Left to Right)
lon1 <- in_lon_n - spacing
lon2 <- in_lon_n 
lon3 <- in_lon_n + spacing

## Get the distance between the points (meters)
line_lon <- rbind(c(lon1,lat1), c(lon2,lat1))
d_lon    <- geosphere::lengthLine(line_lon)

line_lat <- rbind(c(lon1,lat1), c(lon1,lat2))
d_lat    <- geosphere::lengthLine(line_lat)


## Get the nine elevations and resolution distance
point1 <- my_elevation_function(lat=lat1,lon=lon1)
point2 <- my_elevation_function(lat=lat1,lon=lon2)
point3 <- my_elevation_function(lat=lat1,lon=lon3)
point4 <- my_elevation_function(lat=lat2,lon=lon1)
point5 <- my_elevation_function(lat=lat2,lon=lon2)
point6 <- my_elevation_function(lat=lat2,lon=lon3)
point7 <- my_elevation_function(lat=lat3,lon=lon1)
point8 <- my_elevation_function(lat=lat3,lon=lon2)
point9 <- my_elevation_function(lat=lat3,lon=lon3)

point1_e <- point1$elevation
point2_e <- point2$elevation
point3_e <- point3$elevation
point4_e <- point4$elevation
point5_e <- point5$elevation
point6_e <- point6$elevation
point7_e <- point7$elevation
point8_e <- point8$elevation
point9_e <- point9$elevation

points_e <- c(point1_e,point2_e,point3_e,point4_e,point5_e,point6_e,point7_e,point8_e,point9_e)

lats <- c(lat1,lat1,lat1,lat2,lat2,lat2,lat3,lat3,lat3)
lons <- c(lon1,lon2,lon3,lon1,lon2,lon3,lon1,lon2,lon3)

elev_all <- cbind(lats,lons,points_e) ##,points_r)
elev_all <- as.data.frame(elev_all)

c_plot <- ggplot(elev_all,aes(x=lons,y=lats,z=points_e)) +
  geom_contour_filled() +
  geom_point() +
  ggtitle(place)

c_plot

in_resolution <- point5$resolution

ggsave(c_plot, file="site_elevation.png")

```

## Keep working on slope

```{r}

## https://rpubs.com/ncifuentess/592696  == this had very helpful code 

## Create a raster from the nine elevation points (here, it is blank)
xy <- raster(ncols=3,nrows=3, xmn = lon1, xmx = lon3, ymn = lat1, ymx = lat3)

## Get the data into the right form (as a data frame with the longitudes and latitudes as columns)
loc_data    <- cbind(lons,lats)
loc_data_df <- data.frame(loc_data)

## Convert the locations to the raster
xy_raster <- rasterize(loc_data,xy)

## The order of the points had to be changed to accommodate the way the raster stores data
mod_points_e <- c(point7_e,point8_e,point9_e,point4_e,point5_e,point6_e,point1_e,point2_e,point3_e)

## Add the elevations to the raster
values(xy_raster) <- mod_points_e

## Plot the raster for confirmation
plot(xy_raster, main=place)

## Calculate the slope
slope = terrain(xy_raster, opt="slope", unit="degrees", neighbors=8)

## Note that the aspect and flow direction calculations using terrain don't seem to work properly

## Aspect calculation - doing it directly from the data
dzdx <- ((point9_e + 2*point6_e + point3_e) - (point7_e + 2*point4_e + point1_e)) / 8
dzdy <- ((point1_e + 2*point2_e + point3_e) - (point7_e + 2*point8_e + point9_e)) / 8
raw_aspect <- 57.29578 * atan2(dzdy,-dzdx)

## Aspect conversion to degrees
if (raw_aspect < 0){
  aspect <- 90.0 - raw_aspect
  } else if (raw_aspect > 90.0) {
  aspect <- 360.0 - raw_aspect + 90.0
  } else {
  aspect <- 90.0 - raw_aspect
  }

## Calculate the Terrain Roughness Index (TRI)
TRI <- terrain(xy_raster, opt="TRI", neighbors=8)

## The elevation statistics for the analysis area
in_min  <- focal(xy_raster, fun=min, w=matrix(1,nrow=3, ncol=3))
in_max  <- focal(xy_raster, fun=max, w=matrix(1,nrow=3, ncol=3))
in_mean <- focal(xy_raster, fun=mean, w=matrix(1,nrow=3, ncol=3))

in_min  <- in_min[2,2]
in_max  <- in_max[2,2]
in_mean <- in_mean[2,2]

## Prepare data for the water direction calculation
in_min_char   <- as.character(format(round(in_min,  4),nsmall=4))
points_e_char <- as.character(format(round(points_e,4),nsmall=4))

## Find the lowest bin compass point direction
lowest_bin <- grep(in_min_char,points_e_char)
bin_dir    <- c("SW","S","SE","W","C","E","NW","N","NE")
flow_dir   <- bin_dir[lowest_bin]

## Format the values (note: this converts the value to a character string)
round_elevation <- format(round(point5_e,1),nsmall=1)
in_slope        <- format(round(slope[2,2],0),nsmall = 0)
in_TRI          <- format(round(TRI[2,2],1),nsmall = 1)
in_res          <- format(round(in_resolution,0),nsmall = 0)
in_asp          <- as.numeric(format(round(aspect,0),nsmall=0))
in_asp_points   <- cimis_degrees_to_compass(in_asp)
in_mean_char    <- format(round(in_mean,1),nsmall=1)
in_min_char     <- format(round(in_min,1),nsmall=1)
in_max_char     <- format(round(in_max,1),nsmall=1)
d_lon_char      <- format(round(d_lon,1),nsmall=1)
d_lat_char      <- format(round(d_lat,1),nsmall=1)

## Initialize line keeper
line <- NULL

## Report the results
line[1] <- paste0("The elevation of ",place," is ",round_elevation," m.")
line[2] <- paste0("The calculations are based on a local area grid ",d_lon_char," m N-S by ",d_lat_char," m E-W.")
line[3] <- paste0("The site's elevation matrix resolution is ",in_res," m.")
line[4] <- paste0("The slope is ",in_slope ," degrees.")
line[5] <- paste0("The aspect of the slope is ",in_asp," degrees (",in_asp_points,").")
line[6] <- paste0("The site's Terrain Roughness Index (TRI) is ",in_TRI,".")
line[7] <- paste0("The mean elevation of the site's local area is ",in_mean_char," m.")
line[8] <- paste0("The highest point of the site's local area is ",in_max_char," m and the lowest ",in_min_char," m.")
line[9] <- paste0("The water flow direction from the central location is ",flow_dir,".")

line <- data.frame(line)

site_info <- gt(line) |>
  cols_label(line="Site Information")

gtsave(site_info, filename = "site_info_table.png")

```

## Map the location

```{r}
## Library
library(readr)
library(ggmap)
library(ggplot2)
library(grid)
library(pdftools)

## Initialize Google Map key; the is stored in a Project directory.
My_Key <- read_file("P://Hot/Workflow/Workflow/keys/Google_Maps_API_Key.txt")
## Register the Google Maps API Key.
register_google(key = My_Key, account_type = "standard")

lonlat     <- c(in_lon_n,in_lat_n)
map_source <- "google"   ## alternatives: google and stamen

## Site Map
zoom      <- 20
map_type  <- "satellite"
map1 <- get_map(location = lonlat, zoom = zoom, maptype = map_type, source = map_source)

## Plot the site map
m1 <- ggmap(map1) +
  theme_void() +
  theme(panel.border = element_rect(color ="white", fill=NA, size=2)) +
  geom_point(data = loc_data_df, aes(x = lons, y = lats),color="white",size=4) +
  ggtitle(place)

m1

ggsave(m1,file="site_satellite.png")

## Area Map
zoom      <- 16
map_type  <- "hybrid"
map2 <- get_map(location = lonlat, zoom = zoom, maptype = map_type, source = map_source)

## Plot the site map
m2 <- ggmap(map2) +
  theme_void() +
  theme(panel.border = element_rect(color ="white", fill=NA, size=2)) +
  geom_point(data = loc_data_df[5,], aes(x = lons, y = lats),color="white",size=4) +
  ggtitle(place)

m2

ggsave(m2,file="site_hybrid.png")


```

## Aspect the hard way

```{r}

## Aspect calculation
dzdx <- ((point9_e + 2*point6_e + point3_e) - (point7_e + 2*point4_e + point1_e)) / 8
dzdy <- ((point1_e + 2*point2_e + point3_e) - (point7_e + 2*point8_e + point9_e)) / 8
raw_aspect <- 57.29578 * atan2(dzdy,-dzdx)

## Aspect conversion to degrees
if (raw_aspect < 0){
  aspect <- 90.0 - raw_aspect
  } else if (raw_aspect > 90.0) {
  aspect <- 360.0 - raw_aspect + 90.0
  } else {
  aspect <- 90.0 - raw_aspect
  }

## Round the direction and get the compass point equivalent
in_asp <- as.numeric(format(round(aspect,0),nsmall=0))
in_asp_points <- cimis_degrees_to_compass(in_asp)

paste0("Now, it appears that the aspect for ",place," is ",in_asp," degrees (",in_asp_points,").")



```

## QR-Scan: Scan images with QR-Codes

The utility ZBAR is used for the scanning. This is a program loaded on the Windows computer which is invoked with a command line.

Get ZBAR for Windows at: <https://zbar.sourceforge.net/download.html>

Then you install ZBAR on you Windows machine.

Here, this process is adapted by creating a .BAT file with a set of ZBAR commands. This .bat file is submitted to the command line for processing.

> zbarimg --raw temp_qr_file.png

Here is a good way to run this utility:

> zbarimg --xml \*.png \> QR.xml

NOTE

Hint: It might be possible to run the Zbarimg commands using the following general structure:

shell.exec("\\\\network\\path\\file.bat")

The info can be sent to a file where it is picked up back in R.

The xml file has both the file name and the QR Code information!

The XML-decode chunk decodes the XML file contents, returning the file names and, if found, the QR-Code.

This is used with both the Collector Label QR-Codes and the Species QR-Codes.

```{r QR-Scan, echo=FALSE,warning=FALSE,error=FALSE}

## This chunk uses a system call to run the Zbar utility program.


```

```{r XML-Decode, echo=FALSE,warning=FALSE,error=FALSE}

library(stringr)

## Decode the data in the XML file generated by the ZBAR utility.

XML_file <- "QR.xml"

doc = readLines(XML_file)

for (i in seq(1,length(doc),1)){

match <- str_detect(doc[i],".png")
if (match == TRUE) {
  ## print(doc[i])
  str_len <- nchar(doc[i])
  file_name <- substr(doc[i],15,str_len-2)
  ## print(file_name)
}  ## end png match

match <- str_detect(doc[i],"DATA")
if (match == TRUE){
  ## print(doc[i])
  str_len <- nchar(doc[i])
  code_string <- substr(doc[i],51,str_len-19)
  file_line<- paste0(file_name,", ",code_string)
  
  ######
  ## This info could be spooled to a file
  ######
  print(file_line)
  }  ## end DATA match

}  ## end for loop

```
